---
title: "H03: conserved seasonal response at orthologs"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2) })
design <- load_design()
orth <- load_orthologs()
out_dir <- file.path(root, "hypotheses", "H03", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
sp_pairs <- combn(species_all, 2, simplify = FALSE)
alpha <- 0.05
n_perm <- 1000
n_boot <- 1000
set.seed(20261006)
```

## Pre-registered decision rule

> `supported` if rho > 0.2 with permutation p < 0.01 in all three pairs and at
> least three GO-slim terms are enriched (FDR < 0.05) in all three pairwise
> concordant sets. `not supported` if any rho <= 0 or no term is shared.
> Otherwise `inconclusive`.

Model: TP3 vs TP1 from the H02 model (`~ colony + timepoint`, same filter,
same common ortholog set). Spearman rho of apeglm-shrunken LFCs per species
pair; null from 1000 permutations of ortholog assignment. GO-slim enrichment
by Fisher exact test among concordantly DE groups (same sign, FDR < 0.05 in
both species) against all groups tested in both; BH across terms within each
pair.

Operational choices not spelled out in `hypothesis.md` (none changes the rule):

- **Fit.** Same design, filter and common ortholog set as H02, fitted with the
  Wald test so the TP3-vs-TP1 contrast has its own p-value and can be shrunk
  with apeglm (H02 used the LRT; the coefficients are identical). DE calls use
  the Wald BH-adjusted p-value with DESeq2's default independent filtering.
- **Permutation p** is one-sided (the prediction is a positive rho):
  (1 + #{rho_perm >= rho_obs}) / (1 + 1000). Rho also gets a 95% bootstrap CI
  over groups.
- **Enrichment universe** is groups tested in both species of the pair that
  carry at least one GO-slim term; groups with no GO-slim annotation cannot be
  enriched for anything. The test is one-sided (enrichment). Odds ratios with
  95% CIs are reported.
- **"No term is shared"** means zero terms at FDR < 0.05 in all three pairs.

## Data

```{r data}
counts_og <- lapply(setNames(species_all, species_all), function(sp) as_count_matrix(genes_by_ortholog(sp, orthologs = orth)))
prep <- lapply(counts_og, prep_counts, design = design)
common <- Reduce(intersect, lapply(prep, function(p) rownames(p$counts)))

n_tab <- bind_rows(lapply(species_all, function(sp) {
  p <- prep[[sp]]
  tibble(species = sp, samples = ncol(p$counts), colonies = n_distinct(p$info$colony),
         n_TP1 = sum(p$info$timepoint == "TP1"), n_TP3 = sum(p$info$timepoint == "TP3"),
         groups_common = length(common))
}))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab)
```

```{r goslim}
goslim <- orth |>
  filter(group_id %in% common, !is.na(goslim_names), goslim_names != "") |>
  select(group_id, goslim_names) |>
  separate_longer_delim(goslim_names, delim = ";") |>
  mutate(term = trimws(goslim_names)) |>
  filter(term != "") |>
  distinct(group_id, term)
```

`r n_distinct(goslim$group_id)` of `r length(common)` common groups carry at
least one GO-slim term (`r n_distinct(goslim$term)` distinct terms).

## Model

```{r fit}
tp3_vs_tp1 <- function(sp) {
  dds <- fit_timepoint_deseq(prep[[sp]]$counts[common, , drop = FALSE], prep[[sp]]$info, test = "Wald")
  wald <- DESeq2::results(dds, name = "timepoint_TP3_vs_TP1", alpha = alpha)
  shr <- DESeq2::lfcShrink(dds, coef = "timepoint_TP3_vs_TP1", type = "apeglm", quiet = TRUE)
  tibble(species = sp, group_id = rownames(wald), baseMean = wald$baseMean,
         lfc_mle = wald$log2FoldChange, lfc_shrunk = shr$log2FoldChange,
         pvalue = wald$pvalue, padj = wald$padj)
}
res <- bind_rows(lapply(species_all, tp3_vs_tp1))
write_csv(res, file.path(out_dir, "tp3_vs_tp1_per_group.csv"))

knitr::kable(res |> group_by(species) |>
               summarise(tested = sum(!is.na(padj)), de = sum(padj < alpha, na.rm = TRUE),
                         up_in_TP3 = sum(padj < alpha & lfc_shrunk > 0, na.rm = TRUE),
                         down_in_TP3 = sum(padj < alpha & lfc_shrunk < 0, na.rm = TRUE)))
```

```{r fig-scatter, fig.width = 10, fig.height = 3.6}
wide <- res |> select(species, group_id, lfc_shrunk) |>
  pivot_wider(names_from = species, values_from = lfc_shrunk)
pair_long <- bind_rows(lapply(sp_pairs, function(pr)
  tibble(pair = paste(pr, collapse = " vs "), x = wide[[pr[1]]], y = wide[[pr[2]]])))
lim <- quantile(abs(c(pair_long$x, pair_long$y)), 0.999, na.rm = TRUE)
g <- ggplot(pair_long, aes(x, y)) +
  geom_hline(yintercept = 0, colour = "grey70") + geom_vline(xintercept = 0, colour = "grey70") +
  geom_point(alpha = 0.15, size = 0.5) +
  facet_wrap(~ pair) + coord_equal(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
  labs(x = "TP3 vs TP1 shrunken log2 FC, first species", y = "second species") +
  theme_bw()
ggsave(file.path(out_dir, "lfc_scatter.png"), g, width = 10, height = 3.6, dpi = 150)
write_csv(wide, file.path(out_dir, "lfc_scatter_data.csv"))
g
```

## Result against the decision rule

```{r tests}
analyse <- function(r) {
  w <- r |> select(species, group_id, lfc_shrunk) |> pivot_wider(names_from = species, values_from = lfc_shrunk)

  rho_tab <- bind_rows(lapply(sp_pairs, function(pr) {
    x <- w[[pr[1]]]; y <- w[[pr[2]]]
    rho <- cor(x, y, method = "spearman")
    perm <- replicate(n_perm, cor(x, sample(y), method = "spearman"))
    boot <- replicate(n_boot, { i <- sample.int(length(x), replace = TRUE); cor(x[i], y[i], method = "spearman") })
    tibble(pair = paste(pr, collapse = "-"), n_groups = length(x), rho = rho,
           rho_lo = quantile(boot, 0.025), rho_hi = quantile(boot, 0.975),
           perm_p = (1 + sum(perm >= rho)) / (1 + n_perm))
  }))

  enr <- bind_rows(lapply(sp_pairs, function(pr) {
    a <- r |> filter(species == pr[1]) |> select(group_id, padj_a = padj, lfc_a = lfc_shrunk)
    b <- r |> filter(species == pr[2]) |> select(group_id, padj_b = padj, lfc_b = lfc_shrunk)
    ab <- inner_join(a, b, by = "group_id") |>
      filter(!is.na(padj_a), !is.na(padj_b), group_id %in% goslim$group_id) |>
      mutate(concordant = padj_a < alpha & padj_b < alpha & sign(lfc_a) == sign(lfc_b))
    gs <- goslim |> filter(group_id %in% ab$group_id)
    conc <- ab$group_id[ab$concordant]
    bind_rows(lapply(split(gs$group_id, gs$term), function(members) {
      in_term <- ab$group_id %in% members
      in_conc <- ab$group_id %in% conc
      tab <- table(factor(in_term, c(TRUE, FALSE)), factor(in_conc, c(TRUE, FALSE)))
      f <- fisher.test(tab, alternative = "greater")
      f2 <- fisher.test(tab)
      tibble(conc_in_term = tab[1, 1], term_size = sum(in_term), n_concordant = length(conc),
             universe = nrow(ab), odds_ratio = unname(f2$estimate),
             or_lo = f2$conf.int[1], or_hi = f2$conf.int[2], p = f$p.value)
    }), .id = "term") |>
      mutate(pair = paste(pr, collapse = "-"), .before = 1, fdr = p.adjust(p, "BH"))
  }))

  shared <- enr |> group_by(term) |>
    summarise(n_pairs_enriched = sum(fdr < 0.05), .groups = "drop") |>
    filter(n_pairs_enriched == length(sp_pairs)) |> pull(term)

  verdict <-
    if (all(rho_tab$rho > 0.2 & rho_tab$perm_p < 0.01) && length(shared) >= 3) "supported" else
    if (any(rho_tab$rho <= 0) || length(shared) == 0) "not supported" else
    "inconclusive"
  list(rho = rho_tab, enr = enr, shared = shared, verdict = verdict)
}

primary <- analyse(res)
write_csv(primary$rho, file.path(out_dir, "rho_per_pair.csv"))
write_csv(primary$enr, file.path(out_dir, "goslim_enrichment.csv"))
write_csv(tibble(term = primary$shared), file.path(out_dir, "goslim_shared_terms.csv"))
knitr::kable(primary$rho, digits = 3)
```

Concordant sets and GO-slim terms enriched at FDR < 0.05:

```{r enr-table}
knitr::kable(primary$enr |> filter(fdr < 0.05) |> arrange(pair, fdr) |>
               select(pair, term, conc_in_term, term_size, n_concordant, odds_ratio, or_lo, or_hi, fdr),
             digits = 3)
```

Terms enriched in all three pairs (`r length(primary$shared)`):
`r if (length(primary$shared)) paste(primary$shared, collapse = "; ") else "none"`.

**Primary verdict: `r primary$verdict`.**

## Sensitivity checks

### Ortholog groups with avg_identity >= 60

Ortholog assignment errors would pull rho toward zero. Same model fits,
restricted to high-identity groups; DE p-values are BH-adjusted again within
the subset.

```{r sens-ident}
ident_groups <- orth$group_id[!is.na(orth$avg_identity) & orth$avg_identity >= 60]
res_ident <- res |> filter(group_id %in% ident_groups) |>
  group_by(species) |> mutate(padj = p.adjust(pvalue, "BH")) |> ungroup()
sens <- analyse(res_ident)
write_csv(sens$rho, file.path(out_dir, "sens_ident_rho_per_pair.csv"))
write_csv(sens$enr, file.path(out_dir, "sens_ident_goslim_enrichment.csv"))
knitr::kable(sens$rho, digits = 3)
```

Shared terms (`r length(sens$shared)`):
`r if (length(sens$shared)) paste(sens$shared, collapse = "; ") else "none"`.
Verdict: `r sens$verdict`.

```{r summary}
write_csv(tibble(analysis = c("primary", "avg_identity >= 60"),
                 verdict = c(primary$verdict, sens$verdict),
                 n_shared_terms = c(length(primary$shared), length(sens$shared))),
          file.path(out_dir, "verdicts.csv"))
```

## Caveats

- About 10 colonies per species.
- TP1 (January) and TP3 (September) differ in temperature, light, nutrients
  and reproductive timing; a conserved response cannot be attributed to any
  one driver.
- H02 found about 55% of Apul groups DE across the year against about 6% in
  Peve and Ptuh, so concordant sets involving Peve or Ptuh will be small and
  enrichment power low.
- Only part of the ortholog set carries GO-slim annotation, and GO terms are
  inherited from SwissProt hits, which favours conserved genes.
- Groups are not independent (co-regulation), so permutation and Fisher
  p-values are optimistic.
