---
title: "H22: does expression forecast next-timepoint physiology at a held-out site?"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H22", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
traits <- c(AFDW = "Host_AFDW.mg.cm2", calc = "calc.umol.cm2.hr")
tps <- c("TP1", "TP2", "TP3", "TP4")
n_perm <- 200
max_pc <- 20
alpha_en <- 0.5
n_cores <- max(1, parallel::detectCores() - 1)
set.seed(20261008)
```

## Pre-registered decision rule

> - `supported` if dR2 > 0 with BH q < 0.05 for at least one trait in at
>   least two species.
> - `not supported` if dR2 <= 0 for both traits in at least two species.
> - `inconclusive` otherwise.

dR2 = out-of-sample R² of the molecular model (baseline + expression PCs,
elastic net) minus that of the baseline (`y[t+1] ~ y[t] + timepoint`), with
the outer CV holding out one site. Amendment 1 (`hypothesis.md`, before any
analysis) adapts this to the two sites that exist (two folds of 5 colonies;
site is confounded with nutrient), caps the PCs at training rows − 1, fixes
the elastic-net and null details, and keeps traits on the raw scale. p is
from 200 within-transition permutations of y[t+1] among training colonies;
q is BH across the two traits within species.

Implementation notes (not in `hypothesis.md`; none changes the rule):

- Timepoint enters as dummies for the transition (TP2→TP3, TP3→TP4 vs
  TP1→TP2). A dummy that is constant in a training fold is dropped from
  both models in that fold.
- Inner CV uses `cv.glmnet(..., foldid = colony, grouped = FALSE)`, since
  each inner fold has only 2–3 rows.
- Every fold asserts that no test colony is in the training data, and PCA,
  centering and lambda are computed from training rows only.

## Data

```{r data}
phys <- load_physiology() |> select(sample_id, all_of(unname(traits))) |>
  group_by(sample_id) |> summarise(across(everything(), ~ mean(.x, na.rm = TRUE)), .groups = "drop")
orth <- load_orthologs()

prep_species <- function(sp) {
  v <- prep_vst(genes_by_ortholog(sp, orthologs = orth), design)
  info <- design |> filter(sample_id %in% colnames(v$vst)) |> select(sample_id, colony, site, nutrient, timepoint) |>
    left_join(phys, by = "sample_id") |> mutate(t = match(timepoint, tps))
  rows <- lapply(names(traits), function(tr) {
    col <- traits[[tr]]
    a <- info |> transmute(sample_id, colony, site, t, y0 = .data[[col]])
    b <- design |> filter(species == sp) |> select(sample_id, colony, timepoint) |> left_join(phys, by = "sample_id") |>
      transmute(colony, t = match(timepoint, tps) - 1, y1 = .data[[col]])
    a |> inner_join(b, by = c("colony", "t")) |> filter(is.finite(y0), is.finite(y1)) |>
      mutate(trait = tr, interval = factor(t, levels = 1:3), species = sp)
  })
  list(species = sp, vst = v$vst, rows = bind_rows(rows))
}
dat <- lapply(setNames(species_all, species_all), prep_species)

n_tab <- bind_rows(lapply(dat, function(d) d$rows |> group_by(species, trait) |>
  summarise(rows = n(), colonies = n_distinct(colony), Mahana = sum(site == "Mahana"), Manava = sum(site == "Manava"),
            .groups = "drop") |> mutate(ortholog_genes = nrow(d$vst))))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab)
```

## Model

```{r functions}
dummies <- function(r) cbind(i2 = as.numeric(r$t == 2), i3 = as.numeric(r$t == 3))

# Fit both models on `tr` rows, predict `te` rows. `X`: gene x sample matrix.
fit_fold <- function(X, tr, te, y1_train = tr$y1) {
  stopifnot(!any(te$colony %in% tr$colony))
  D_tr <- dummies(tr); D_te <- dummies(te)
  keep <- apply(D_tr, 2, function(z) length(unique(z)) > 1)
  D_tr <- D_tr[, keep, drop = FALSE]; D_te <- D_te[, keep, drop = FALSE]
  B_tr <- cbind(y0 = tr$y0, D_tr); B_te <- cbind(y0 = te$y0, D_te)
  base <- lm.fit(cbind(1, B_tr), y1_train)
  pred_base <- drop(cbind(1, B_te) %*% base$coefficients)
  Xtr <- t(X[, tr$sample_id, drop = FALSE]); mu <- colMeans(Xtr)
  k <- min(max_pc, nrow(Xtr) - 1)
  pc <- prcomp(sweep(Xtr, 2, mu), center = FALSE, rank. = k)
  S_tr <- pc$x; S_te <- sweep(t(X[, te$sample_id, drop = FALSE]), 2, mu) %*% pc$rotation
  Z_tr <- cbind(B_tr, S_tr); Z_te <- cbind(B_te, S_te)
  pf <- c(rep(0, ncol(B_tr)), rep(1, ncol(S_tr)))
  foldid <- as.integer(factor(tr$colony))
  cv <- glmnet::cv.glmnet(Z_tr, y1_train, alpha = alpha_en, penalty.factor = pf, foldid = foldid,
                          grouped = FALSE, standardize = TRUE)
  pred_mol <- drop(predict(cv, Z_te, s = "lambda.min"))
  tibble(row = te$row, y1 = te$y1, pred_base = pred_base, pred_mol = pred_mol, k = k,
         n_pc_selected = sum(as.matrix(coef(cv, s = "lambda.min"))[-(1:(ncol(B_tr) + 1)), 1] != 0))
}

r2_pooled <- function(p) {
  sst <- sum((p$y1 - mean(p$y1))^2)
  c(R2_base = 1 - sum((p$y1 - p$pred_base)^2) / sst, R2_mol = 1 - sum((p$y1 - p$pred_mol)^2) / sst)
}

# Outer CV over a grouping column (`site` or `colony`); optional permutation of training y1.
run_cv <- function(X, rows, by = "site", permute = FALSE) {
  rows <- rows |> mutate(row = row_number())
  folds <- unique(rows[[by]])
  preds <- bind_rows(lapply(folds, function(f) {
    tr <- rows[rows[[by]] != f, ]; te <- rows[rows[[by]] == f, ]
    y1 <- tr$y1
    if (permute) for (g in unique(tr$t)) { i <- which(tr$t == g); y1[i] <- y1[i][sample.int(length(i))] }
    fit_fold(X, tr, te, y1)
  }))
  list(preds = preds, r2 = r2_pooled(preds))
}

evaluate <- function(d, trait, by = "site", rows_filter = function(r) r) {
  rows <- rows_filter(d$rows |> filter(.data$trait == !!trait))
  obs <- run_cv(d$vst, rows, by)
  null <- unlist(parallel::mclapply(seq_len(n_perm), function(i) {
    set.seed(20261008 + i)
    r <- run_cv(d$vst, rows, by, permute = TRUE)$r2
    r[["R2_mol"]] - r[["R2_base"]]
  }, mc.cores = n_cores))
  dR2 <- obs$r2[["R2_mol"]] - obs$r2[["R2_base"]]
  list(summary = tibble(species = d$species, trait = trait, cv = by, rows = nrow(rows),
                        R2_base = obs$r2[["R2_base"]], R2_mol = obs$r2[["R2_mol"]], dR2 = dR2,
                        null_dR2_median = median(null), null_dR2_95 = quantile(null, 0.95),
                        p = (1 + sum(null >= dR2)) / (1 + length(null)),
                        mean_pcs_selected = mean(obs$preds$n_pc_selected), pcs_available = max(obs$preds$k)),
       preds = obs$preds |> mutate(species = d$species, trait = trait, cv = by), null = null)
}

decide <- function(s) {
  s <- s |> group_by(species) |> mutate(q = p.adjust(p, "BH")) |> ungroup()
  per <- s |> group_by(species) |>
    summarise(any_pass = any(dR2 > 0 & q < 0.05), both_nonpos = all(dR2 <= 0), .groups = "drop")
  verdict <- if (sum(per$any_pass) >= 2) "supported" else if (sum(per$both_nonpos) >= 2) "not supported" else "inconclusive"
  list(per_trait = s, per_species = per, verdict = verdict)
}
```

```{r primary}
res <- lapply(species_all, function(sp) lapply(names(traits), function(tr) evaluate(dat[[sp]], tr, "site")))
res <- unlist(res, recursive = FALSE)
primary <- decide(bind_rows(lapply(res, `[[`, "summary")))
write_csv(primary$per_trait, file.path(out_dir, "dr2_summary.csv"))
write_csv(bind_rows(lapply(res, `[[`, "preds")), file.path(out_dir, "predictions.csv"))
write_csv(bind_rows(lapply(res, function(r) tibble(species = r$summary$species, trait = r$summary$trait, null_dR2 = r$null))),
          file.path(out_dir, "null_dr2.csv"))
knitr::kable(primary$per_trait, digits = 3)
```

**Primary verdict: `r primary$verdict`.**

```{r fig, fig.width = 9, fig.height = 5}
fp <- bind_rows(lapply(res, `[[`, "preds")) |> left_join(bind_rows(lapply(dat, `[[`, "rows")) |>
  group_by(species, trait) |> mutate(row = row_number()) |> ungroup() |> select(species, trait, row, site), by = c("species", "trait", "row")) |>
  pivot_longer(c(pred_base, pred_mol), names_to = "model", values_to = "pred") |>
  mutate(model = recode(model, pred_base = "baseline", pred_mol = "baseline + expression"))
write_csv(fp, file.path(out_dir, "figure_data.csv"))
g <- ggplot(fp, aes(pred, y1, colour = model)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70") +
  geom_point(alpha = 0.7, size = 1.6) +
  facet_wrap(trait ~ species, scales = "free", nrow = 2) +
  scale_colour_manual(values = c(baseline = "grey45", `baseline + expression` = "#1b6ca8")) +
  labs(x = "Predicted y[t+1] (held-out site)", y = "Observed y[t+1]", colour = NULL) +
  theme_bw() + theme(legend.position = "bottom")
ggsave(file.path(out_dir, "observed_vs_predicted.png"), g, width = 9, height = 5.5, dpi = 150)
g
```

## Sensitivity checks

```{r sensitivity}
sens_loco <- decide(bind_rows(lapply(species_all, function(sp) bind_rows(lapply(names(traits), function(tr)
  evaluate(dat[[sp]], tr, "colony")$summary)))))
sens_no34 <- decide(bind_rows(lapply(species_all, function(sp) bind_rows(lapply(names(traits), function(tr)
  evaluate(dat[[sp]], tr, "site", rows_filter = function(r) r |> filter(t != 3))$summary)))))
write_csv(sens_loco$per_trait, file.path(out_dir, "sens_leave_one_colony_out.csv"))
write_csv(sens_no34$per_trait, file.path(out_dir, "sens_drop_tp3_tp4.csv"))
knitr::kable(bind_rows(sens_loco$per_trait |> mutate(check = "leave one colony out"),
                       sens_no34$per_trait |> mutate(check = "TP3->TP4 dropped")) |>
               select(check, species, trait, rows, R2_base, R2_mol, dR2, p, q), digits = 3)
```

## Secondary: cross-species transfer (descriptive)

Train on all rows of one species, test on another; ortholog genes and both
y columns z-scored within species. No permutation null and no decision.

```{r transfer}
common <- Reduce(intersect, lapply(dat, function(d) rownames(d$vst)))
zg <- lapply(dat, function(d) t(scale(t(d$vst[common, ]))))
zrows <- lapply(dat, function(d) d$rows |> group_by(trait) |>
  mutate(m = mean(c(y0, y1)), s = sd(c(y0, y1)), y0 = (y0 - m) / s, y1 = (y1 - m) / s) |> ungroup())
transfer <- bind_rows(lapply(species_all, function(a) bind_rows(lapply(setdiff(species_all, a), function(b)
  bind_rows(lapply(names(traits), function(tr) {
    tr_rows <- zrows[[a]] |> filter(trait == tr) |> mutate(row = row_number())
    te_rows <- zrows[[b]] |> filter(trait == tr) |> mutate(row = row_number(), colony = paste0("test_", colony))
    X <- cbind(zg[[a]], zg[[b]])
    p <- fit_fold(X, tr_rows, te_rows)
    r <- r2_pooled(p)
    tibble(train = a, test = b, trait = tr, genes = length(common), R2_base = r[["R2_base"]], R2_mol = r[["R2_mol"]],
           dR2 = r[["R2_mol"]] - r[["R2_base"]])
  }))))))
write_csv(transfer, file.path(out_dir, "cross_species_transfer.csv"))
knitr::kable(transfer, digits = 3)
```

```{r summary}
verdicts <- tibble(analysis = c("primary (leave one site out)", "leave one colony out", "TP3->TP4 dropped"),
                   verdict = c(primary$verdict, sens_loco$verdict, sens_no34$verdict))
write_csv(verdicts, file.path(out_dir, "verdicts.csv"))
knitr::kable(verdicts)
```

## Caveats

- About 10 colonies per species: 5 per site, so each fold trains on 5
  colonies (about 13–15 rows).
- Site is confounded with nutrient; "novel conditions" means the other
  site and nutrient regime, and nothing more.
- No independent cohort exists; held-out site and held-out species are
  proxies.
- Raw-scale traits; AFDW includes negative values from upstream.
