---
title: "Hypothesis status board"
---

```{r}
root <- ".."
source(file.path(root, "R", "status.R"))
board <- hypothesis_board(file.path(root, "hypotheses"))
lock  <- yaml::read_yaml(file.path(root, "config", "upstream.lock.yml"))
```

Built against MOSAiC commit `r substr(lock$mosaic_commit, 1, 7)`; upstream lock last updated `r lock$lock_updated`.
Rules for adding a hypothesis are in [CLAUDE.md](https://github.com/urol-e5/MOSAiC-hypotheses/blob/main/CLAUDE.md).

```{r}
library(gt)
status_colors <- c(planned = "#e9ecef", running = "#fff3cd", supported = "#d1e7dd",
                   `not supported` = "#f8d7da", inconclusive = "#e2e3e5", withdrawn = "#dee2e6")
board |>
  dplyr::mutate(layers = vapply(layers, paste, "", collapse = ", "),
                species = vapply(species, paste, "", collapse = ", "),
                depends_on = vapply(depends_on, function(x) if (length(x)) paste(x, collapse = ", ") else "", "")) |>
  dplyr::select(id, tier, title, status, layers, species, depends_on) |>
  gt() |>
  tab_header(title = "Hypotheses") |>
  cols_label(id = "ID", tier = "Tier", title = "Claim", status = "Status",
             layers = "Layers", species = "Species", depends_on = "Depends on") |>
  data_color(columns = status, fn = function(x) unname(status_colors[x])) |>
  tab_options(table.font.size = px(13))
```

## Tiers

1. **Variance structure.** Colony vs season vs site. Everything downstream depends on these.
2. **Regulatory layers.** Methylation, miRNA, lncRNA.
3. **Phenotype and environment.** Physiology, symbionts, thermal history, nutrients.
4. **Integration.** Multi-omics factors and energetics.
