---
title: "Data QC and design overview"
---

```{r setup}
root <- ".."
source(file.path(root, "R", "load.R"))
suppressPackageStartupMessages({ library(dplyr); library(tidyr); library(ggplot2); library(gt); library(matrixStats) })
design <- readr::read_csv(file.path(root, "config", "design.csv"), show_col_types = FALSE)
derived <- function(n) readr::read_csv(file.path(root, "data", "derived", paste0(n, ".csv")), show_col_types = FALSE, progress = FALSE)
lock <- yaml::read_yaml(file.path(root, "config", "upstream.lock.yml"))
sp_names <- c(Apul = "A. pulchra", Peve = "P. evermanni", Ptuh = "P. tuahiniensis")
```

Harmonized from MOSAiC commit `r substr(lock$mosaic_commit, 1, 7)`; lock updated `r lock$lock_updated`.
Every table here is derived from `config/design.csv` and `data/derived/`, which are what every hypothesis reads.

## Samples per species, timepoint, and layer

```{r}
layer_cols <- grep("^has_", names(design), value = TRUE)
design |>
  group_by(species, timepoint) |>
  summarise(colonies = n_distinct(colony), across(all_of(layer_cols), sum), .groups = "drop") |>
  rename_with(~ sub("^has_", "", .x)) |>
  gt(groupname_col = "species") |>
  tab_header(title = "Sample counts by layer") |>
  tab_options(table.font.size = px(13))
```

## Colony coverage

Which colonies have which timepoints in the gene-expression layer. Gaps matter for every repeated-measures model.

```{r}
#| fig-height: 7
design |>
  filter(has_genes) |>
  mutate(species = sp_names[species]) |>
  ggplot(aes(timepoint, colony)) +
  geom_tile(aes(fill = site), colour = "white") +
  facet_wrap(~ species, scales = "free_y", ncol = 3) +
  labs(x = NULL, y = NULL, fill = "Site", title = "Gene-expression samples per colony") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")
```

## Layer overlap

Number of samples shared between each pair of layers (all species pooled). The diagonal is the layer's own n.

```{r}
L <- as.matrix(design[, layer_cols]); colnames(L) <- sub("^has_", "", layer_cols)
ov <- crossprod(L * 1L)
as.data.frame(ov) |> tibble::rownames_to_column("layer") |> gt() |>
  tab_header(title = "Pairwise sample overlap") |> tab_options(table.font.size = px(13))
```

Samples present in **every** molecular layer (genes, mirna, lncrna, cpg, metabolomics, lipidomics), per species:

```{r}
mol <- c("has_genes", "has_mirna", "has_lncrna", "has_cpg", "has_metabolomics", "has_lipidomics")
design |> filter(if_all(all_of(mol))) |> dplyr::count(species, timepoint) |>
  pivot_wider(names_from = timepoint, values_from = n, values_fill = 0) |> gt()
```

## Site and nutrient balance

```{r}
design |> distinct(species, colony, site, nutrient) |>
  dplyr::count(species, site, nutrient) |>
  pivot_wider(names_from = species, values_from = n, values_fill = 0) |>
  gt() |> tab_header(title = "Colonies per site x nutrient")
```

## Gene-expression library sizes and PCA

```{r}
#| fig-height: 4
libs <- bind_rows(lapply(names(sp_names), function(sp) {
  g <- derived(paste0("genes_", sp))
  tibble(species = sp, sample_id = setdiff(names(g), "feature_id"),
         libsize = colSums(as_count_matrix(g)))
})) |> left_join(design, by = c("species", "sample_id"))
ggplot(libs, aes(timepoint, libsize / 1e6, colour = site)) +
  geom_jitter(width = 0.15, size = 2) +
  facet_wrap(~ species, labeller = labeller(species = sp_names)) +
  labs(y = "Assigned reads (millions)", x = NULL, title = "Gene count library sizes") +
  theme_minimal(base_size = 11)
```

```{r}
#| fig-height: 4
pca_df <- bind_rows(lapply(names(sp_names), function(sp) {
  m <- as_count_matrix(derived(paste0("genes_", sp)))
  keep <- rowSums(m >= 10) >= 0.25 * ncol(m)
  m <- m[keep, ]
  l <- log2(t(t(m) / colSums(m) * 1e6) + 1)
  v <- rowVars(l); l <- l[order(v, decreasing = TRUE)[1:min(2000, nrow(l))], ]
  p <- prcomp(t(l), scale. = FALSE)
  ve <- round(100 * p$sdev[1:2]^2 / sum(p$sdev^2))
  tibble(species = sp, sample_id = rownames(p$x), PC1 = p$x[, 1], PC2 = p$x[, 2],
         lab = sprintf("PC1 %d%% / PC2 %d%%", ve[1], ve[2]))
})) |> left_join(design, by = c("species", "sample_id"))
ggplot(pca_df, aes(PC1, PC2, colour = timepoint, shape = site)) +
  geom_point(size = 2.5) +
  facet_wrap(~ species + lab, scales = "free") +
  labs(title = "PCA of log-CPM, top 2000 variable genes") +
  theme_minimal(base_size = 11)
```

Colonies are connected across timepoints below; tight clusters per colony indicate colony-dominated variance (the H01 question).

```{r}
#| fig-height: 4
ggplot(pca_df, aes(PC1, PC2, group = colony, colour = colony)) +
  geom_path(alpha = 0.6) + geom_point(aes(shape = timepoint), size = 2) +
  facet_wrap(~ species, scales = "free") +
  theme_minimal(base_size = 11) + theme(legend.position = "none") +
  labs(title = "Same PCA, points joined by colony")
```

## Ortholog coverage

```{r}
og <- derived("orthologs_three_way")
tibble(species = names(sp_names),
       orthologs_in_counts = sapply(names(sp_names), function(sp) {
         key <- c(Apul = "apul_gene", Peve = "peve_gene", Ptuh = "ptuh_gene")[[sp]]
         sum(og[[key]] %in% derived(paste0("genes_", sp))$feature_id)
       }),
       total = nrow(og),
       with_protein_name = sum(!is.na(og$protein_name) & og$protein_name != "")) |>
  gt() |> tab_header(title = "Three-way ortholog groups resolvable in each count matrix")
```

## Temperature history before each timepoint

Daily mean per site with the nominal timepoint dates (mid-month; decision D-006).

```{r}
#| fig-height: 4
temp <- derived("temperature_daily")
tp <- tibble(timepoint = paste0("TP", 1:4), date = as.Date(c("2020-01-15", "2020-03-15", "2020-09-15", "2020-11-15")))
ggplot(temp, aes(day, mean_c, colour = site)) +
  geom_ribbon(aes(ymin = min_c, ymax = max_c, fill = site), alpha = 0.15, colour = NA) +
  geom_line() +
  geom_vline(data = tp, aes(xintercept = date), linetype = "dotted") +
  geom_text(data = tp, aes(x = date, y = 31.5, label = timepoint), inherit.aes = FALSE, size = 3) +
  labs(x = NULL, y = "Daily temperature (C)", title = "Logger record by site") +
  theme_minimal(base_size = 11)
```

```{r}
temp |>
  tidyr::crossing(tp) |>
  filter(day > date - 30, day <= date) |>
  group_by(site, timepoint) |>
  summarise(days = n(), mean_30d = round(mean(mean_c), 2), max_30d = round(max(max_c), 2), .groups = "drop") |>
  pivot_wider(names_from = timepoint, values_from = c(mean_30d, max_30d, days)) |>
  gt() |> tab_header(title = "30-day window before each nominal timepoint date")
```

## Upstream files

```{r}
tibble::as_tibble(do.call(rbind, lapply(lock$files, as.data.frame))) |>
  mutate(MB = round(as.numeric(bytes) / 1e6, 2)) |> select(id, MB, md5, fetched_on) |>
  gt() |> tab_options(table.font.size = px(12))
```
