---
title: "H15: persistence of seasonal methylation vs expression changes"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
source(file.path(root, "R", "methylation.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H15", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
n_boot <- 2000
sel_frac <- 0.10
set.seed(20261007)
```

## Pre-registered decision rule

> - `supported` if dP > 0 with 95% CI excluding 0 in at least two species,
>   and in those species P_GBM > mean P_T.
> - `not supported` if the dP point estimate is <= 0 in at least two species.
> - `inconclusive` otherwise, including when b31 for GBM has a CI that
>   includes 0 (no detectable seasonal methylation shift to persist) in two
>   or more species.

P = persistence = mean over colonies of b41_c / mean over colonies of b31_c,
where b31_c and b41_c are the through-origin slopes of colony c's
(x_TP3 − x_TP1) and (x_TP4 − x_TP1) on D_g, the leave-one-colony-out (LOCO)
mean of (x_TP3 − x_TP1), over that colony's top 10% of genes by |D_g|.
dP = P_GBM − P_expr. CIs by colony bootstrap (2000 resamples), with LOCO
selection redone inside each resample.

Implementation notes (not in `hypothesis.md`; none changes the rule):

- The rule's no-shift `inconclusive` clause is checked before
  `supported` or `not supported`: if the GBM b31 CI includes 0 in two or
  more species there is no shift whose persistence can be measured, and the verdict is
  `inconclusive` regardless of dP.
- In a bootstrap resample a colony drawn twice is not its own LOCO
  reference: D for a draw is the mean over draws of *other* original
  colonies.
- Bootstrap p-values (two-sided, 2 × the smaller tail) are Holm-corrected
  across species for reporting only.
- Temperature: mean of daily site means over the 30 days before each
  nominal date (D-006). P_T = (T4 − T1)/(T3 − T1) per site, averaged over
  the species' colonies (so weighted by colonies per site).

## Data

```{r data}
build_layers <- function(sp, min_cpg = 5) {
  cpg <- as_count_matrix(load_cpg(sp))
  both <- design |> filter(species == sp, has_cpg %in% TRUE, has_genes %in% TRUE, sample_id %in% colnames(cpg))
  ex <- prep_vst(load_genes(sp) |> select(feature_id, all_of(both$sample_id)), design)
  trip <- both |> filter(timepoint %in% c("TP1", "TP3", "TP4"), sample_id %in% colnames(ex$vst)) |>
    group_by(colony) |> filter(n_distinct(timepoint) == 3) |> ungroup() |> arrange(colony, timepoint)
  gb <- gene_body_methylation(cpg[, trip$sample_id, drop = FALSE], load_gene_coords(sp), min_cpg = min_cpg)
  genes <- intersect(rownames(gb$gbm), rownames(ex$vst))
  cols <- unique(trip$colony)
  pick <- function(tp) trip$sample_id[trip$timepoint == tp][match(cols, trip$colony[trip$timepoint == tp])]
  s1 <- pick("TP1"); s3 <- pick("TP3"); s4 <- pick("TP4")
  mk <- function(m) {
    d31 <- m[genes, s3, drop = FALSE] - m[genes, s1, drop = FALSE]
    d41 <- m[genes, s4, drop = FALSE] - m[genes, s1, drop = FALSE]
    colnames(d31) <- colnames(d41) <- cols
    list(d31 = d31, d41 = d41)
  }
  list(species = sp, colonies = cols, sites = trip$site[match(cols, trip$colony)],
       gbm = mk(gb$gbm), expr = mk(ex$vst), n_cpg = gb$n_cpg[genes],
       n = tibble(species = sp, colonies_TP1_TP3_TP4 = length(cols), samples_both_layers = nrow(both),
                  genes_used = length(genes)))
}

lay <- lapply(setNames(species_all, species_all), build_layers)
n_tab <- bind_rows(lapply(lay, `[[`, "n"))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab, format.args = list(big.mark = ","))
```

Colonies used (both layers at TP1, TP3 and TP4):

```{r colonies}
knitr::kable(bind_rows(lapply(lay, function(d) tibble(species = d$species, colonies = paste(d$colonies, collapse = ", ")))))
```

### Temperature persistence

```{r temperature}
temp <- load_temperature_daily()
tp_dates <- design |> distinct(timepoint, tp_date) |> filter(timepoint %in% c("TP1", "TP3", "TP4"))
t30 <- tidyr::crossing(site = unique(temp$site), tp_dates) |>
  rowwise() |>
  mutate(T30 = mean(temp$mean_c[temp$site == site & temp$day < tp_date & temp$day >= tp_date - 30]),
         days = sum(temp$site == site & temp$day < tp_date & temp$day >= tp_date - 30)) |>
  ungroup()
pt_site <- t30 |> select(site, timepoint, T30) |> pivot_wider(names_from = timepoint, values_from = T30) |>
  mutate(P_T = (TP4 - TP1) / (TP3 - TP1))
write_csv(t30, file.path(out_dir, "temperature_30d.csv"))
knitr::kable(pt_site, digits = 3)
P_T_species <- sapply(lay, function(d) mean(pt_site$P_T[match(d$sites, pt_site$site)]))
```

## Model

```{r functions}
# Persistence for one layer. d31, d41: gene x colony change matrices.
# `ids` gives the original colony identity of each column (differs from
# colnames only in bootstrap resamples), so LOCO excludes every copy of a colony.
persistence <- function(d31, d41, ids = colnames(d31), frac = sel_frac) {
  k <- ncol(d31)
  b <- t(sapply(seq_len(k), function(j) {
    other <- ids != ids[j]
    D <- rowMeans(d31[, other, drop = FALSE])
    sel <- order(abs(D), decreasing = TRUE)[seq_len(ceiling(frac * length(D)))]
    D <- D[sel]
    c(b31 = sum(D * d31[sel, j]) / sum(D^2), b41 = sum(D * d41[sel, j]) / sum(D^2),
      flip31 = mean(sign(d31[sel, j]) != sign(D)), flip41 = mean(sign(d41[sel, j]) != sign(D)))
  }))
  c(P = mean(b[, "b41"]) / mean(b[, "b31"]), b31 = mean(b[, "b31"]), b41 = mean(b[, "b41"]),
    flip31 = mean(b[, "flip31"]), flip41 = mean(b[, "flip41"]))
}

analyse <- function(d, genes = rownames(d$gbm$d31), frac = sel_frac) {
  g <- d$gbm; e <- d$expr
  est <- function(cols, ids) {
    pg <- persistence(g$d31[genes, cols, drop = FALSE], g$d41[genes, cols, drop = FALSE], ids, frac)
    pe <- persistence(e$d31[genes, cols, drop = FALSE], e$d41[genes, cols, drop = FALSE], ids, frac)
    c(P_gbm = pg[["P"]], P_expr = pe[["P"]], dP = pg[["P"]] - pe[["P"]],
      b31_gbm = pg[["b31"]], b41_gbm = pg[["b41"]], b31_expr = pe[["b31"]], b41_expr = pe[["b41"]],
      flip31_gbm = pg[["flip31"]], flip41_gbm = pg[["flip41"]],
      flip31_expr = pe[["flip31"]], flip41_expr = pe[["flip41"]])
  }
  cols <- d$colonies
  obs <- est(cols, cols)
  boot <- t(replicate(n_boot, {
    i <- sample.int(length(cols), replace = TRUE)
    if (length(unique(i)) < 3) return(rep(NA_real_, length(obs)))
    est(cols[i], cols[i])
  }))
  colnames(boot) <- names(obs)
  ci <- function(v) quantile(boot[, v], c(0.025, 0.975), na.rm = TRUE)
  p2 <- function(v) min(1, 2 * min(mean(boot[, v] <= 0, na.rm = TRUE), mean(boot[, v] >= 0, na.rm = TRUE)))
  list(
    summary = tibble(species = d$species, colonies = length(cols), genes_selected = ceiling(frac * length(genes)),
                     genes = length(genes),
                     P_gbm = obs[["P_gbm"]], P_gbm_lo = ci("P_gbm")[1], P_gbm_hi = ci("P_gbm")[2],
                     P_expr = obs[["P_expr"]], P_expr_lo = ci("P_expr")[1], P_expr_hi = ci("P_expr")[2],
                     dP = obs[["dP"]], dP_lo = ci("dP")[1], dP_hi = ci("dP")[2], dP_boot_p = p2("dP"),
                     b31_gbm = obs[["b31_gbm"]], b31_gbm_lo = ci("b31_gbm")[1], b31_gbm_hi = ci("b31_gbm")[2],
                     b31_expr = obs[["b31_expr"]], b31_expr_lo = ci("b31_expr")[1], b31_expr_hi = ci("b31_expr")[2],
                     P_T = P_T_species[[d$species]],
                     n_boot_valid = sum(!is.na(boot[, "dP"]))),
    rebound = tibble(species = d$species, layer = c("gbm", "expr"),
                     opposite_sign_TP3 = c(obs[["flip31_gbm"]], obs[["flip31_expr"]]),
                     opposite_sign_TP4 = c(obs[["flip41_gbm"]], obs[["flip41_expr"]])),
    boot = as_tibble(boot) |> mutate(species = d$species))
}

decide <- function(s) {
  s <- s |> mutate(dP_boot_p_holm = p.adjust(dP_boot_p, "holm"),
                   shift_gbm = b31_gbm_lo > 0 | b31_gbm_hi < 0,
                   passes = dP_lo > 0 & P_gbm > P_T)
  verdict <- if (sum(!s$shift_gbm) >= 2) "inconclusive"
    else if (sum(s$passes) >= 2) "supported"
    else if (sum(s$dP <= 0) >= 2) "not supported"
    else "inconclusive"
  list(per_species = s, verdict = verdict)
}
```

```{r fit}
res <- lapply(lay, analyse)
primary <- decide(bind_rows(lapply(res, `[[`, "summary")))
write_csv(primary$per_species, file.path(out_dir, "persistence_summary.csv"))
write_csv(bind_rows(lapply(res, `[[`, "boot")), file.path(out_dir, "persistence_bootstrap.csv"))
knitr::kable(primary$per_species |> select(species, colonies, genes, P_gbm, P_gbm_lo, P_gbm_hi, P_expr, P_expr_lo, P_expr_hi,
                                           dP, dP_lo, dP_hi, dP_boot_p_holm, b31_gbm_lo, b31_gbm_hi, P_T, passes), digits = 3)
```

**Primary verdict: `r primary$verdict`.**

```{r fig-trajectory, fig.width = 9, fig.height = 3.5}
traj <- primary$per_species |>
  transmute(species, gbm = P_gbm, gbm_lo = P_gbm_lo, gbm_hi = P_gbm_hi,
            expr = P_expr, expr_lo = P_expr_lo, expr_hi = P_expr_hi, temperature = P_T) |>
  pivot_longer(-species, names_to = c("layer", "stat"), names_pattern = "([a-z]+)_?(lo|hi)?") |>
  mutate(stat = ifelse(stat == "", "est", stat)) |>
  pivot_wider(names_from = stat, values_from = value)
traj <- bind_rows(
  traj |> mutate(timepoint = "TP4"),
  traj |> mutate(timepoint = "TP3", est = 1, lo = NA, hi = NA),
  traj |> mutate(timepoint = "TP1", est = 0, lo = NA, hi = NA)) |>
  mutate(timepoint = factor(timepoint, c("TP1", "TP3", "TP4")),
         layer = factor(layer, c("gbm", "expr", "temperature"), c("gene-body methylation", "expression", "temperature")))
write_csv(traj, file.path(out_dir, "trajectory_normalized.csv"))
g <- ggplot(traj, aes(timepoint, est, colour = layer, group = layer)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_line(position = position_dodge(0.25)) + geom_point(position = position_dodge(0.25)) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.15, position = position_dodge(0.25), na.rm = TRUE) +
  facet_wrap(~ species) +
  scale_colour_manual(values = c("#1b6ca8", "#d1495b", "grey40")) +
  coord_cartesian(ylim = c(-3, 3)) +
  labs(x = NULL, y = "Fraction of TP1-to-TP3 shift present", colour = NULL) +
  theme_bw() + theme(legend.position = "bottom")
ggsave(file.path(out_dir, "persistence_trajectory.png"), g, width = 9, height = 3.8, dpi = 150)
g
```

The TP4 point is P; 0 is full reversion to TP1, 1 is no reversion. Error bars
are 95% colony-bootstrap intervals; the y-axis is clipped at ±3, so wider
intervals run off the panel (full values in `persistence_summary.csv`).

### Rebound (secondary, descriptive)

Fraction of a colony's selected genes whose change from TP1 has the opposite
sign to the LOCO shift D, averaged over colonies. At TP3 this is the noise
floor (the shift is defined from the other colonies); a TP4 value above it
means some genes overshoot past baseline. Relates to Q4; no decision.

```{r rebound}
reb <- bind_rows(lapply(res, `[[`, "rebound"))
write_csv(reb, file.path(out_dir, "rebound.csv"))
knitr::kable(reb, digits = 3)
```

## Sensitivity checks

```{r sensitivity}
sens_run <- function(label, gene_fun = function(d) rownames(d$gbm$d31), frac = sel_frac) {
  s <- bind_rows(lapply(lay, function(d) analyse(d, gene_fun(d), frac)$summary))
  dec <- decide(s)
  write_csv(dec$per_species, file.path(out_dir, paste0("sens_", label, "_summary.csv")))
  list(label = label, dec = dec)
}

orth <- load_orthologs()
orth_ids <- list(Apul = orth$apul_gene, Peve = orth$peve_gene, Ptuh = orth$ptuh_gene)

sens <- list(
  sens_run("mincpg20", function(d) names(d$n_cpg)[d$n_cpg >= 20]),
  sens_run("top5", frac = 0.05),
  sens_run("top20", frac = 0.20),
  sens_run("orthologs", function(d) intersect(rownames(d$gbm$d31), orth_ids[[d$species]])))
sens_tab <- bind_rows(lapply(sens, function(x) x$dec$per_species |>
  transmute(check = x$label, species, genes, P_gbm, P_expr, dP, dP_lo, dP_hi, passes)))
knitr::kable(sens_tab, digits = 3)
```

```{r summary}
verdicts <- tibble(analysis = c("primary (top 10%, >= 5 CpGs)", sapply(sens, `[[`, "label")),
                   verdict = c(primary$verdict, sapply(sens, function(x) x$dec$verdict)))
write_csv(verdicts, file.path(out_dir, "verdicts.csv"))
knitr::kable(verdicts)
```

## Caveats

- Colonies with both layers at TP1, TP3 and TP4:
  `r paste(n_tab$species, n_tab$colonies_TP1_TP3_TP4, sep = " ", collapse = ", ")`.
  About 10 colonies per species before this restriction.
- P is a ratio; when b31 is close to 0 for a layer, P and its interval are
  unstable. Read P only where the b31 interval excludes 0.
- TP4 is two months after TP3; P describes persistence over that interval
  only, and TP3→TP4 also spans the *Acropora* spawning window.
- CpG density differs about 20-fold across species (D-008), so gene sets
  differ in size and composition.
