---
title: "H06: seasonal methylation change vs expression change"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
source(file.path(root, "R", "methylation.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2); library(lme4) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H06", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
n_perm <- 1000
n_boot <- 1000
n_sim <- 200
dm_frac <- 0.10
set.seed(20261006)
```

## Pre-registered decision rule

> `supported` if rho > 0 with permutation p < 0.05 in at least two species and
> the mixed-model GBM slope is positive with CI excluding zero in those species.
> `not supported` if rho <= 0 in all three. Otherwise `inconclusive`.

**Amendment 1 (2026-10-07, before any analysis; see `hypothesis.md`).** The
originally stated null (shuffle colony labels for a colony-averaged rho) is
degenerate. As amended, "rho" in the rule is the per-colony coupling
statistic T = mean over colonies of Spearman(dGBM_c, dExpr_c) across genes,
with a null that pairs each colony's methylation change with another colony's
expression change (1000 permutations, one-sided). DM genes are the top 10% by
|colony-averaged dGBM|; the mixed model `expr ~ gbm + (1|colony) + (1|gene)`
is fit on TP1 and TP3 samples of paired colonies for the DM set. The
colony-averaged rho is descriptive only.

No correction across species: the rule counts species passing at p < 0.05,
as pre-registered.

Implementation notes (not in `hypothesis.md`; none changes the rule):

- Because ranks are fixed per colony, the full colony-by-colony matrix of
  rho(dGBM_c, dExpr_c') is computed once; T and every permuted T are means of
  its entries.
- The mixed-model slope uses a Wald 95% CI, in vst units per 10 percentage
  points of GBM.
- **Power** (pre-registered sensitivity): at each species' observed n, add a
  coupling of strength b to the observed expression deltas,
  dExpr*_c = dExpr_pi(c) + b · sd(dExpr) · z(dGBM_c), with pi a random
  colony permutation so that no real coupling remains, and record how often
  the permutation test gives p < 0.05 (`r n_sim` simulations per b).

## Data

```{r data}
build_deltas <- function(sp, coords) {
  cpg <- as_count_matrix(load_cpg(sp))
  both <- design |> filter(species == sp, has_cpg %in% TRUE, has_genes %in% TRUE, sample_id %in% colnames(cpg))
  ex <- prep_vst(load_genes(sp) |> select(feature_id, all_of(both$sample_id)), design)
  paired <- both |> filter(timepoint %in% c("TP1", "TP3"), sample_id %in% colnames(ex$vst)) |>
    group_by(colony) |> filter(n_distinct(timepoint) == 2) |> ungroup() |> arrange(colony, timepoint)
  gb <- gene_body_methylation(cpg[, paired$sample_id, drop = FALSE], coords, min_cpg = 5)
  genes <- intersect(rownames(gb$gbm), rownames(ex$vst))
  cols <- unique(paired$colony)
  s1 <- paired$sample_id[paired$timepoint == "TP1"][match(cols, paired$colony[paired$timepoint == "TP1"])]
  s3 <- paired$sample_id[paired$timepoint == "TP3"][match(cols, paired$colony[paired$timepoint == "TP3"])]
  dG <- gb$gbm[genes, s3, drop = FALSE] - gb$gbm[genes, s1, drop = FALSE]
  dE <- ex$vst[genes, s3, drop = FALSE] - ex$vst[genes, s1, drop = FALSE]
  colnames(dG) <- colnames(dE) <- cols
  list(species = sp, dG = dG, dE = dE, gbm = gb$gbm[genes, paired$sample_id, drop = FALSE],
       vst = ex$vst[genes, paired$sample_id, drop = FALSE], paired = paired,
       n = tibble(species = sp, colonies_paired = length(cols), samples_both_layers = nrow(both),
                  genes_used = length(genes)))
}

gb_data <- lapply(setNames(species_all, species_all), function(sp) build_deltas(sp, load_gene_coords(sp)))
n_tab <- bind_rows(lapply(gb_data, `[[`, "n"))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab, format.args = list(big.mark = ","))
```

Paired colonies (both layers at TP1 and TP3):

```{r paired}
knitr::kable(bind_rows(lapply(gb_data, function(d) tibble(species = d$species, colonies = paste(colnames(d$dG), collapse = ", ")))))
```

## Model

```{r functions}
# Colony x colony matrix of Spearman rho(dGBM of row colony, dExpr of column colony).
coupling_matrix <- function(dG, dE) {
  rG <- apply(dG, 2, rank); rE <- apply(dE, 2, rank)
  cor(rG, rE)
}

perm_test <- function(R, B = n_perm) {
  k <- nrow(R)
  T_obs <- mean(diag(R))
  T_perm <- replicate(B, { p <- sample.int(k); mean(R[cbind(seq_len(k), p)]) })
  list(T = T_obs, p = (1 + sum(T_perm >= T_obs)) / (1 + B), T_perm = T_perm)
}

mixed_slope <- function(d, genes) {
  long <- tibble(
    gene = rep(genes, times = ncol(d$gbm)),
    sample_id = rep(colnames(d$gbm), each = length(genes)),
    gbm10 = as.vector(d$gbm[genes, , drop = FALSE]) / 10,
    expr = as.vector(d$vst[genes, colnames(d$gbm), drop = FALSE])) |>
    left_join(d$paired |> select(sample_id, colony), by = "sample_id")
  fit <- lmer(expr ~ gbm10 + (1 | colony) + (1 | gene), data = long)
  ci <- confint(fit, parm = "gbm10", method = "Wald")
  tibble(n_obs = nrow(long), slope = unname(fixef(fit)["gbm10"]), slope_lo = ci[1], slope_hi = ci[2])
}

analyse <- function(d) {
  R <- coupling_matrix(d$dG, d$dE)
  pt <- perm_test(R)
  mG <- rowMeans(d$dG); mE <- rowMeans(d$dE)
  avg_rho <- cor(mG, mE, method = "spearman")
  avg_boot <- replicate(n_boot, { i <- sample.int(length(mG), replace = TRUE); cor(mG[i], mE[i], method = "spearman") })
  dm <- names(sort(abs(mG), decreasing = TRUE))[seq_len(ceiling(dm_frac * length(mG)))]
  R_dm <- coupling_matrix(d$dG[dm, , drop = FALSE], d$dE[dm, , drop = FALSE])
  pt_dm <- perm_test(R_dm)
  list(
    summary = tibble(species = d$species, colonies = ncol(d$dG), genes = nrow(d$dG),
                     T = pt$T, perm_p = pt$p,
                     T_null_median = median(pt$T_perm),
                     avg_rho = avg_rho, avg_rho_lo = quantile(avg_boot, 0.025), avg_rho_hi = quantile(avg_boot, 0.975),
                     dm_genes = length(dm), T_dm = pt_dm$T, perm_p_dm = pt_dm$p) |>
      bind_cols(mixed_slope(d, dm)),
    per_colony = tibble(species = d$species, colony = colnames(d$dG), rho_c = diag(R)),
    R = R)
}

decide <- function(s) {
  s <- s |> mutate(passes = T > 0 & perm_p < 0.05 & slope > 0 & slope_lo > 0)
  verdict <- if (sum(s$passes) >= 2) "supported" else if (all(s$T <= 0)) "not supported" else "inconclusive"
  list(per_species = s, verdict = verdict)
}
```

```{r fit}
res <- lapply(gb_data, analyse)
primary <- decide(bind_rows(lapply(res, `[[`, "summary")))
write_csv(primary$per_species, file.path(out_dir, "coupling_summary.csv"))
write_csv(bind_rows(lapply(res, `[[`, "per_colony")), file.path(out_dir, "coupling_per_colony.csv"))
write_csv(bind_rows(lapply(gb_data, function(d)
  tibble(species = d$species, feature_id = rownames(d$dG), mean_dgbm = rowMeans(d$dG), mean_dexpr = rowMeans(d$dE)))),
  file.path(out_dir, "deltas_per_gene.csv"))
knitr::kable(primary$per_species, digits = 4)
```

**Primary verdict: `r primary$verdict`.**

```{r fig-coupling, fig.width = 9, fig.height = 3.5}
R_long <- bind_rows(lapply(res, function(r) {
  R <- r$R
  tibble(species = r$summary$species, meth_colony = rep(rownames(R), ncol(R)),
         expr_colony = rep(colnames(R), each = nrow(R)), rho = as.vector(R),
         same_colony = meth_colony == expr_colony)
}))
write_csv(R_long, file.path(out_dir, "coupling_matrix.csv"))
g <- ggplot(R_long, aes(same_colony, rho)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1.2) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, colour = "black") +
  facet_wrap(~ species) +
  scale_x_discrete(labels = c(`FALSE` = "different colony\n(null pairs)", `TRUE` = "same colony")) +
  labs(x = NULL, y = "Spearman rho, dGBM vs dExpr across genes") +
  theme_bw()
ggsave(file.path(out_dir, "coupling_same_vs_other_colony.png"), g, width = 9, height = 3.5, dpi = 150)
g
```

## Sensitivity checks

### Power at the observed n

```{r power}
b_grid <- c(0.01, 0.02, 0.03, 0.05, 0.1)
power_one <- function(d, b) {
  k <- ncol(d$dG)
  zG <- scale(d$dG)
  sims <- replicate(n_sim, {
    p0 <- sample.int(k)
    dE_null <- d$dE[, p0, drop = FALSE]
    dE_sim <- dE_null + b * sd(as.vector(d$dE)) * zG
    R <- coupling_matrix(d$dG, dE_sim)
    pt <- perm_test(R, B = 199)
    c(T = pt$T, sig = pt$p < 0.05)
  })
  tibble(species = d$species, b = b, mean_T = mean(sims["T", ]), power = mean(sims["sig", ]))
}
power_tab <- bind_rows(lapply(gb_data, function(d) bind_rows(lapply(b_grid, power_one, d = d))))
write_csv(power_tab, file.path(out_dir, "power.csv"))
knitr::kable(power_tab, digits = 3)
```

`b` is the added coupling in SD units of expression change per SD of
methylation change; `mean_T` is the per-colony coupling it produces on top of
shuffled (uncoupled) data. Power uses 199 permutations per simulation.

### 2 kb upstream windows

```{r sens-upstream}
up_data <- lapply(setNames(species_all, species_all), function(sp) build_deltas(sp, upstream_coords(load_gene_coords(sp), 2000)))
res_up <- lapply(up_data, analyse)
sens_up <- decide(bind_rows(lapply(res_up, `[[`, "summary")))
write_csv(sens_up$per_species, file.path(out_dir, "sens_upstream2kb_summary.csv"))
knitr::kable(sens_up$per_species, digits = 4)
```

Verdict for upstream windows: `r sens_up$verdict`.

```{r summary}
write_csv(tibble(analysis = c("gene body (primary)", "2 kb upstream"), verdict = c(primary$verdict, sens_up$verdict)),
          file.path(out_dir, "verdicts.csv"))
```

## Caveats

- Paired colonies: `r paste(n_tab$species, n_tab$colonies_paired, sep = " ", collapse = ", ")`.
  The permutation null has at most k! distinct pairings and is only as
  informative as the number of colonies allows; see the power table.
- About 10 colonies per species before pairing.
- Per-colony deltas from one TP1 and one TP3 sample carry the full sampling
  noise of both layers; colony-level GBM from 10x CpGs is noisy at the few-%
  methylation typical of these genes.
- CpG density differs about 20-fold across species (D-008), so the gene sets
  differ in size and composition.
- TP1 and TP3 differ in temperature, light, nutrients and reproductive
  timing; coupling, if present, is not attributable to temperature.
