---
title: "H17: lncRNA–neighbor gene coupling vs an expression-matched null"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
source(file.path(root, "R", "neighbors.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H17", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
n_null <- 1000      # null draws of the median
n_cand <- 100       # candidate null partners per cis pair
n_boot <- 2000      # bootstrap over pairs for the median-r CI
max_dist <- 100000
bins <- c(0, 2000, 10000, 50000, 100000)
bin_labels <- c("0-2 kb", "2-10 kb", "10-50 kb", "50-100 kb")
set.seed(20261007)
```

## Pre-registered decision rule

> - `supported` if dr > 0 with null p < 0.01 and dr >= 0.05 in all three
>   species, and the distance-decay rho is negative in at least two.
> - `not supported` if dr < 0.02 or null p >= 0.05 in at least two species.
> - `inconclusive` otherwise.

dr = median Pearson r over cis pairs (lncRNA and gene on the same scaffold,
not overlapping, nearest edges <= 10 kb apart) minus the mean of 1000 null
medians, each from one expression-matched partner per cis pair drawn from
other scaffolds. Both layers are vst'd on the same samples and residualized
on colony. Amendment 1 (`hypothesis.md`, before any analysis) fixes the
null construction and records that lncRNA strand is unknown upstream (D-014):
the orientation split is dropped and the read-through check removes all pairs
within 2 kb.

Implementation notes (not in `hypothesis.md`; none changes the rule):

- Distance is the gap between the nearest ends of the two intervals in bp;
  pairs with gap <= 0 overlap. Bins are (0, 2], (2, 10], (10, 50] and
  (50, 100] kb.
- Mean-vst deciles are computed over all filtered genes of the species
  after vst and before residualization.
- Distance-decay rho is the Spearman correlation of the four bin medians
  with bin rank (1–4); with four points it is a sign check, as
  pre-registered.
- The decision uses one-sided null p (is cis r larger than the null). The
  `not supported` clause "null p >= 0.05" uses the same p.

## Data

```{r data}
residualize <- function(m, info, terms) {
  X <- model.matrix(as.formula(paste("~", paste(terms, collapse = " + "))), data = info)
  t(qr.resid(qr(X), t(m)))
}
zrows <- function(m) { m <- m - rowMeans(m); m / sqrt(rowSums(m^2)) }   # rows with unit norm: r = dot product

prep_species <- function(sp) {
  ids <- design |> filter(species == sp, has_genes %in% TRUE, has_lncrna %in% TRUE) |> pull(sample_id)
  g <- prep_vst(load_genes(sp) |> select(feature_id, any_of(ids)), design)
  l <- prep_vst(load_lncrna(sp) |> select(feature_id, any_of(ids)), design)
  ids <- intersect(colnames(g$vst), colnames(l$vst))
  G <- g$vst[, ids]; L <- l$vst[, ids]
  info <- design_for(G, design) |> mutate(colony = factor(colony), timepoint = factor(timepoint))
  gc <- load_gene_coords(sp) |> filter(feature_id %in% rownames(G))
  lc <- lncrna_coords(sp) |> filter(feature_id %in% rownames(L))
  decile <- dplyr::ntile(rowMeans(G), 10); names(decile) <- rownames(G)
  list(species = sp, G = G, L = L, info = info, gc = gc, lc = lc, decile = decile,
       n = tibble(species = sp, samples = length(ids), colonies = n_distinct(info$colony),
                  df_after_colony = length(ids) - n_distinct(info$colony),
                  lncrna = nrow(L), genes = nrow(G),
                  lncrna_with_coords = nrow(lc), genes_with_coords = nrow(gc)))
}

dat <- lapply(setNames(species_all, species_all), prep_species)
n_tab <- bind_rows(lapply(dat, `[[`, "n"))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab, format.args = list(big.mark = ","))
```

## Model

```{r functions}
# Pearson r for row pairs (ia of A, ib of B), A and B already z-normed by row.
pair_r <- function(A, B, ia, ib, chunk = 2e5) {
  out <- numeric(length(ia))
  for (s in seq(1, length(ia), by = chunk)) {
    k <- s:min(length(ia), s + chunk - 1)
    out[k] <- rowSums(A[ia[k], , drop = FALSE] * B[ib[k], , drop = FALSE])
  }
  out
}

# Observed r for pairs, plus null medians for the cis (<= 10 kb, non-overlap) set.
# `A`, `B`: z-normed matrices of the two layers; pairs$a in A, pairs$b in B.
# Null partners for pair p come from B rows on other scaffolds in the decile of pairs$b[p].
run_set <- function(d, A, B, pairs, b_coords, cis_filter) {
  pairs$r <- pair_r(A, B, match(pairs$a, rownames(A)), match(pairs$b, rownames(B)))
  cis <- pairs |> filter(!overlap, cis_filter(gap))
  b_chrom <- setNames(b_coords$chrom, b_coords$feature_id)[rownames(B)]
  b_dec <- d$decile[rownames(B)]
  pools <- split(seq_len(nrow(B)), b_dec)
  # candidates: n_cand per cis pair, same decile, other scaffold
  cand <- matrix(NA_integer_, nrow(cis), n_cand)
  dec_cis <- d$decile[cis$b]
  for (p in seq_len(nrow(cis))) {
    pool <- pools[[as.character(dec_cis[p])]]
    pool <- pool[b_chrom[pool] != cis$chrom[p]]
    cand[p, ] <- pool[sample.int(length(pool), n_cand)]
  }
  ia <- rep(match(cis$a, rownames(A)), times = n_cand)
  cand_r <- matrix(pair_r(A, B, ia, as.vector(cand)), nrow(cis), n_cand)
  null_med <- replicate(n_null, median(cand_r[cbind(seq_len(nrow(cis)), sample.int(n_cand, nrow(cis), replace = TRUE))]))
  obs_med <- median(cis$r)
  boot_med <- replicate(n_boot, median(sample(cis$r, replace = TRUE)))
  list(pairs = pairs, cis = cis, null_med = null_med,
       summary = tibble(cis_pairs = nrow(cis), cis_lncrna = n_distinct(cis$a),
                        median_r = obs_med, median_r_lo = quantile(boot_med, 0.025), median_r_hi = quantile(boot_med, 0.975),
                        null_median_mean = mean(null_med),
                        dr = obs_med - mean(null_med),
                        dr_lo = obs_med - quantile(null_med, 0.975), dr_hi = obs_med - quantile(null_med, 0.025),
                        p = (1 + sum(null_med >= obs_med)) / (1 + n_null)))
}

bin_table <- function(pairs) {
  pairs |> filter(!overlap) |> group_by(bin) |>
    summarise(pairs = n(), median_r = median(r), mean_r = mean(r), .groups = "drop")
}

analyse <- function(d, terms = "colony", cis_filter = function(g) g <= 10000) {
  L <- zrows(residualize(d$L, d$info, terms)); G <- zrows(residualize(d$G, d$info, terms))
  pairs <- neighbor_pairs(d$lc, d$gc)
  rs <- run_set(d, L, G, pairs, d$gc, cis_filter)
  bt <- bin_table(rs$pairs)
  decay <- suppressWarnings(cor(seq_len(nrow(bt)), bt$median_r, method = "spearman"))
  ov <- rs$pairs |> filter(overlap)
  list(summary = tibble(species = d$species) |> bind_cols(rs$summary) |>
         mutate(decay_rho = decay, overlap_pairs = nrow(ov), overlap_median_r = if (nrow(ov)) median(ov$r) else NA_real_),
       bins = bt |> mutate(species = d$species, .before = 1),
       null_med = tibble(species = d$species, null_median = rs$null_med),
       pairs = rs$pairs |> mutate(species = d$species, .before = 1))
}

decide <- function(s) {
  s <- s |> mutate(passes = dr > 0 & p < 0.01 & dr >= 0.05, fails = dr < 0.02 | p >= 0.05)
  verdict <- if (all(s$passes) && sum(s$decay_rho < 0, na.rm = TRUE) >= 2) "supported"
    else if (sum(s$fails) >= 2) "not supported"
    else "inconclusive"
  list(per_species = s, verdict = verdict)
}
```

```{r fit}
res <- lapply(dat, analyse)
primary <- decide(bind_rows(lapply(res, `[[`, "summary")))
bins_tab <- bind_rows(lapply(res, `[[`, "bins"))
write_csv(primary$per_species, file.path(out_dir, "cis_vs_null_summary.csv"))
write_csv(bins_tab, file.path(out_dir, "r_by_distance_bin.csv"))
write_csv(bind_rows(lapply(res, `[[`, "null_med")), file.path(out_dir, "null_medians.csv"))
knitr::kable(primary$per_species |> select(species, cis_pairs, cis_lncrna, median_r, null_median_mean, dr, dr_lo, dr_hi, p,
                                           decay_rho, passes, fails), digits = 4)
```

**Primary verdict: `r primary$verdict`.**

Pairs and median r per distance bin (non-overlapping pairs):

```{r bins}
knitr::kable(bins_tab, digits = 4, format.args = list(big.mark = ","))
```

Overlapping lncRNA–gene pairs (excluded from the test; shared reads can
create correlation):

```{r overlap}
knitr::kable(primary$per_species |> select(species, overlap_pairs, overlap_median_r), digits = 4)
```

```{r fig-bins, fig.width = 9, fig.height = 3.8}
nulls <- primary$per_species |> select(species, null_median_mean)
g <- ggplot(bins_tab, aes(bin, median_r)) +
  geom_hline(data = nulls, aes(yintercept = null_median_mean), linetype = 2, colour = "grey40") +
  geom_point(aes(size = pairs), colour = "#1b6ca8") + geom_line(aes(group = 1), colour = "#1b6ca8") +
  facet_wrap(~ species) + scale_size_area(max_size = 4) +
  labs(x = "lncRNA-gene distance (nearest edges)", y = "Median r (colony-residualized)", size = "Pairs") +
  theme_bw() + theme(axis.text.x = element_text(angle = 30, hjust = 1))
ggsave(file.path(out_dir, "r_by_distance.png"), g, width = 9, height = 3.8, dpi = 150)
g
```

Dashed line: mean null median (expression-matched genes on other scaffolds,
for the <= 10 kb set). Data in `r_by_distance_bin.csv` and
`cis_vs_null_summary.csv`.

**Orientation split (Model step 6):** not estimable. Upstream lncRNA
features carry no strand (all `+`, D-014).

## Sensitivity checks

### Timepoint also residualized

```{r sens-tp}
res_tp <- lapply(dat, analyse, terms = c("colony", "timepoint"))
sens_tp <- decide(bind_rows(lapply(res_tp, `[[`, "summary")))
write_csv(sens_tp$per_species, file.path(out_dir, "sens_timepoint_resid_summary.csv"))
write_csv(bind_rows(lapply(res_tp, `[[`, "bins")), file.path(out_dir, "sens_timepoint_resid_bins.csv"))
knitr::kable(sens_tp$per_species |> select(species, cis_pairs, median_r, null_median_mean, dr, dr_lo, dr_hi, p, decay_rho), digits = 4)
```

### Read-through: drop all pairs within 2 kb

```{r sens-rt}
res_rt <- lapply(dat, analyse, cis_filter = function(g) g > 2000 & g <= 10000)
sens_rt <- decide(bind_rows(lapply(res_rt, `[[`, "summary")))
write_csv(sens_rt$per_species, file.path(out_dir, "sens_no_2kb_summary.csv"))
knitr::kable(sens_rt$per_species |> select(species, cis_pairs, median_r, null_median_mean, dr, dr_lo, dr_hi, p, decay_rho), digits = 4)
```

### Gene–gene neighbor baseline

Same distances, same null construction, with gene pairs in place of
lncRNA–gene pairs (each unordered pair once; the null partner is matched to
the second gene's decile). A lncRNA-specific cis effect should exceed this.

```{r sens-gg}
gg_one <- function(d) {
  G <- zrows(residualize(d$G, d$info, "colony"))
  pairs <- neighbor_pairs(d$gc, d$gc, same = TRUE)
  rs <- run_set(d, G, G, pairs, d$gc, function(g) g <= 10000)
  list(summary = tibble(species = d$species) |> bind_cols(rs$summary),
       bins = bin_table(rs$pairs) |> mutate(species = d$species, .before = 1))
}
gg <- lapply(dat, gg_one)
gg_tab <- bind_rows(lapply(gg, `[[`, "summary")) |>
  left_join(primary$per_species |> select(species, dr_lncrna = dr), by = "species") |>
  mutate(dr_lncrna_minus_gene = dr_lncrna - dr)
write_csv(gg_tab, file.path(out_dir, "sens_gene_gene_summary.csv"))
write_csv(bind_rows(lapply(gg, `[[`, "bins")), file.path(out_dir, "sens_gene_gene_bins.csv"))
knitr::kable(gg_tab |> select(species, cis_pairs, median_r, null_median_mean, dr, dr_lo, dr_hi, p, dr_lncrna, dr_lncrna_minus_gene), digits = 4)
```

```{r fig-gg, fig.width = 9, fig.height = 3.8}
cmp <- bind_rows(bins_tab |> mutate(pair_type = "lncRNA-gene"),
                 bind_rows(lapply(gg, `[[`, "bins")) |> mutate(pair_type = "gene-gene"))
write_csv(cmp, file.path(out_dir, "lncrna_vs_gene_gene_bins.csv"))
g2 <- ggplot(cmp, aes(bin, median_r, colour = pair_type, group = pair_type)) +
  geom_line() + geom_point() + facet_wrap(~ species) +
  scale_colour_manual(values = c(`lncRNA-gene` = "#1b6ca8", `gene-gene` = "grey45")) +
  labs(x = "Distance (nearest edges)", y = "Median r (colony-residualized)", colour = NULL) +
  theme_bw() + theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "bottom")
ggsave(file.path(out_dir, "lncrna_vs_gene_gene.png"), g2, width = 9, height = 4, dpi = 150)
g2
```

```{r summary}
verdicts <- tibble(analysis = c("primary (colony-residualized, <= 10 kb)", "timepoint also residualized", "pairs within 2 kb dropped"),
                   verdict = c(primary$verdict, sens_tp$verdict, sens_rt$verdict))
write_csv(verdicts, file.path(out_dir, "verdicts.csv"))
knitr::kable(verdicts)
```

## Caveats

- About 10 colonies per species; samples and residual df in the first table.
- Pairs are not independent: one lncRNA can pair with several genes, and one
  gene with several lncRNAs. The null keeps that structure (one partner per
  observed pair), but the bootstrap CI on the median resamples pairs as if
  independent and is optimistic.
- Both layers come from the same libraries (see H08). Null partners are
  measured in the same samples, so per-sample technical structure is in both
  the observed and the null medians and cancels in dr; it does not cancel for
  local (scaffold-level) effects, which is what the gene–gene baseline
  addresses.
- Correlation is not function. A positive dr would show cis co-variation
  above background, not that the lncRNA regulates the gene.
- lncRNA strand unknown (D-014).
