---
title: "H05: gene-body methylation and expression level and stability"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
source(file.path(root, "R", "methylation.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2); library(splines) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H05", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
n_boot <- 1000
set.seed(20261006)
```

## Pre-registered decision rule

> `supported` if rho(GBM, mean expression) > 0.2 and the GBM coefficient on CV
> is negative with a CI excluding zero, in all three species. `not supported` if
> either sign is reversed with a CI excluding zero in any species. Otherwise
> `inconclusive`.

Model: per species, per gene: mean GBM (%), mean vst expression, CV of
per-timepoint colony-mean expression. Spearman rho(GBM, mean expression); then
`lm(cv ~ ns(mean_expr, 3) + gbm)` with the standardized GBM coefficient and
95% CI. Genes need >= 5 CpGs in the gene body; samples as in H04 (both
`has_cpg` and `has_genes`). Decisions use per-species 95% CIs; no p-value
enters the rule, so no multiple-testing correction is applied.

Operational choices not spelled out in `hypothesis.md` (none changes the rule):

- **Gene body** is the GFF `gene` feature span (start to end, both strands);
  a CpG inside overlapping genes counts toward each. CpG positions are parsed
  from the CpG ids; an off-by-one in their coordinate base cannot matter at
  gene-body scale.
- **GBM per gene** is the mean over samples of the per-sample mean percent
  methylation across the gene's CpGs.
- **Expression** is vst on the intersected samples with the H01 filter; genes
  failing that filter are dropped. **CV** = sd / mean of the four
  per-timepoint means, each the mean over colonies of that timepoint's vst
  values (vst is a log scale, so CV is of log-expression, as pre-registered).
- **Standardized coefficient**: `cv` and `gbm` are z-scored before the fit, so
  the GBM coefficient is in SD of CV per SD of GBM.
- **Rho CI** is a 95% bootstrap over genes. "Reversed with a CI excluding
  zero" means rho < 0 with the whole CI below zero, or the GBM coefficient
  > 0 with the whole CI above zero.

## Data

```{r data}
build_species <- function(sp, min_cpg) {
  cpg <- as_count_matrix(load_cpg(sp))
  ids <- intersect(colnames(cpg), design$sample_id[design$has_cpg %in% TRUE & design$has_genes %in% TRUE])
  ex <- prep_vst(load_genes(sp) |> select(feature_id, all_of(ids)), design)
  ids <- intersect(ids, colnames(ex$vst))
  gb <- gene_body_methylation(cpg[, ids, drop = FALSE], load_gene_coords(sp), min_cpg = min_cpg)
  info <- design_for(ex$vst[, ids, drop = FALSE], design)
  v <- ex$vst[, ids, drop = FALSE]
  tp_means <- sapply(split(seq_along(ids), info$timepoint), function(j) rowMeans(v[, j, drop = FALSE]))
  genes <- intersect(rownames(v), rownames(gb$gbm))
  list(
    per_gene = tibble(species = sp, feature_id = genes,
                      n_cpg = unname(gb$n_cpg[genes]),
                      gbm = rowMeans(gb$gbm[genes, , drop = FALSE]),
                      mean_expr = rowMeans(v[genes, , drop = FALSE]),
                      cv = apply(tp_means[genes, , drop = FALSE], 1, sd) / rowMeans(tp_means[genes, , drop = FALSE])),
    n = tibble(species = sp, samples = length(ids), colonies = n_distinct(info$colony),
               cpg_sites = nrow(cpg), genes_in_gff = nrow(load_gene_coords(sp)),
               genes_with_min_cpg = nrow(gb$gbm), genes_expressed = nrow(v), genes_used = length(genes),
               median_cpg_per_gene = median(gb$n_cpg[genes])))
}

dat <- lapply(species_all, build_species, min_cpg = 5)
per_gene <- bind_rows(lapply(dat, `[[`, "per_gene"))
n_tab <- bind_rows(lapply(dat, `[[`, "n"))
write_csv(per_gene, file.path(out_dir, "per_gene.csv"))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab, format.args = list(big.mark = ","))
```

CpG sites differ about 20-fold across species (decision D-008), so the number
of genes reaching 5 CpGs does too.

## Model

```{r model}
analyse <- function(pg) {
  bind_rows(lapply(split(pg, pg$species), function(d) {
    rho <- cor(d$gbm, d$mean_expr, method = "spearman")
    boot <- replicate(n_boot, { i <- sample.int(nrow(d), replace = TRUE); cor(d$gbm[i], d$mean_expr[i], method = "spearman") })
    d <- d |> mutate(cv_z = as.numeric(scale(cv)), gbm_z = as.numeric(scale(gbm)))
    fit <- lm(cv_z ~ ns(mean_expr, 3) + gbm_z, data = d)
    ci <- confint(fit)["gbm_z", ]
    tibble(species = d$species[1], n_genes = nrow(d),
           rho = rho, rho_lo = quantile(boot, 0.025), rho_hi = quantile(boot, 0.975),
           beta_gbm = unname(coef(fit)["gbm_z"]), beta_lo = ci[1], beta_hi = ci[2],
           beta_p = summary(fit)$coefficients["gbm_z", 4])
  }))
}

decide <- function(est) {
  est <- est |> mutate(
    supports = rho > 0.2 & beta_gbm < 0 & beta_hi < 0,
    reverses = rho_hi < 0 | beta_lo > 0)
  list(per_species = est,
       verdict = if (all(est$supports)) "supported" else if (any(est$reverses)) "not supported" else "inconclusive")
}

primary <- decide(analyse(per_gene))
write_csv(primary$per_species, file.path(out_dir, "estimates.csv"))
knitr::kable(primary$per_species, digits = 3)
```

**Primary verdict: `r primary$verdict`.**

### GBM deciles

```{r deciles, fig.width = 9, fig.height = 5}
dec <- per_gene |>
  group_by(species) |>
  mutate(gbm_decile = ntile(gbm, 10)) |>
  group_by(species, gbm_decile) |>
  summarise(gbm_median = median(gbm), n = n(),
            mean_expr_median = median(mean_expr), mean_expr_q25 = quantile(mean_expr, 0.25), mean_expr_q75 = quantile(mean_expr, 0.75),
            cv_median = median(cv), cv_q25 = quantile(cv, 0.25), cv_q75 = quantile(cv, 0.75), .groups = "drop")
write_csv(dec, file.path(out_dir, "gbm_deciles.csv"))

dec_long <- bind_rows(
  dec |> transmute(species, gbm_decile, panel = "mean vst expression", med = mean_expr_median, lo = mean_expr_q25, hi = mean_expr_q75),
  dec |> transmute(species, gbm_decile, panel = "CV across timepoints", med = cv_median, lo = cv_q25, hi = cv_q75))
g <- ggplot(dec_long, aes(gbm_decile, med)) +
  geom_pointrange(aes(ymin = lo, ymax = hi), size = 0.3) +
  facet_grid(panel ~ species, scales = "free_y") +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Gene-body methylation decile (1 = lowest)", y = "Median (bars: IQR)") +
  theme_bw()
ggsave(file.path(out_dir, "gbm_deciles.png"), g, width = 9, height = 5, dpi = 150)
g
```

## Sensitivity checks

### Lower CpG threshold (>= 2 CpGs per gene)

```{r sens-mincpg}
dat2 <- lapply(species_all, build_species, min_cpg = 2)
per_gene2 <- bind_rows(lapply(dat2, `[[`, "per_gene"))
sens <- decide(analyse(per_gene2))
write_csv(bind_rows(lapply(dat2, `[[`, "n")), file.path(out_dir, "sens_mincpg2_n.csv"))
write_csv(sens$per_species, file.path(out_dir, "sens_mincpg2_estimates.csv"))
knitr::kable(sens$per_species, digits = 3)
```

Verdict with >= 2 CpGs: `r sens$verdict`.

```{r summary}
write_csv(tibble(analysis = c("primary (>= 5 CpGs)", ">= 2 CpGs"), verdict = c(primary$verdict, sens$verdict)),
          file.path(out_dir, "verdicts.csv"))
```

## Caveats

- About 10 colonies per species; GBM and CV are per-gene summaries over
  `r paste(n_tab$samples, collapse = " / ")` samples (Apul / Peve / Ptuh).
- CV over four timepoint means is noisy, and on the vst (log) scale it is
  tied to mean level, which is why mean expression enters as a spline.
- The upstream 10x-in-all-samples CpG filter is not the same stringency
  across species (D-008), so GBM is estimated from very different CpG
  densities; Ptuh is mapped to *P. meandrina* and Peve has a fragmented
  assembly (N50 0.17 Mb), which also affects gene models.
- Gene bodies include introns and UTRs where annotated; GFF quality differs
  by species.
