---
title: "H08: lncRNA-mRNA co-expression against a permutation null"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H08", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
thresholds <- c(0.8, 0.9, 0.95)
set.seed(20261006)

# Compute tier is raven (README): each species needs about 800 passes over
# 1e8 to 3e8 lncRNA x mRNA correlations (about 25 s per Apul pass with
# reference BLAS). Permutations run in parallel with forked workers; each
# needs about 200 MB. H08_WORKERS sets the count (default cores - 1).
# H08_N_PERM (default 200, as pre-registered) may be lowered for a dry run.
n_workers <- as.integer(Sys.getenv("H08_WORKERS", max(1, parallel::detectCores() - 1)))
n_perm <- as.integer(Sys.getenv("H08_N_PERM", "200"))
```

## Pre-registered decision rule

> `supported` if the observed count exceeds the null 95th percentile at all
> three thresholds in all three species. `not supported` if the observed count
> falls inside the null at |r| >= 0.9 in any species. Otherwise `inconclusive`.

Model: Pearson r on vst values for all lncRNA x mRNA pairs, samples with both
layers, features filtered as in H01. Null: 200 permutations of lncRNA sample
order; a within-timepoint permutation is also reported. Count of pairs with
|r| >= 0.8, 0.9, 0.95, observed vs null. No correction across thresholds or
species: the rule is stated per threshold and species.

Operational choices not spelled out in `hypothesis.md` (none changes the rule):

- **Which null decides.** The full permutation, as the Model section names
  it. The within-timepoint permutation (lncRNA labels shuffled only among
  samples from the same timepoint, so the seasonal signal both layers share
  is kept in the null) and the colony-residualized analysis are reported
  with their own verdicts.
- **"Exceeds"** is strictly greater than the null 95th percentile; "falls
  inside the null" is less than or equal to it.
- **vst** is computed separately for each layer (blind, design `~ 1`).
- **Residualizing on colony** removes colony means from each feature of both
  layers (least squares on a colony factor) before z-scoring; the same nulls
  are then applied to the residualized lncRNA matrix.
- **Permutation indices** are drawn once, before any correlations, so the run
  is reproducible regardless of the number of workers. Empirical
  p = (1 + #{null >= observed}) / (1 + n_perm).

## Data

```{r data}
prep_layers <- function(sp) {
  ids <- design$sample_id[design$species == sp & design$has_lncrna %in% TRUE & design$has_genes %in% TRUE]
  g <- prep_vst(load_genes(sp) |> select(feature_id, any_of(ids)), design)
  l <- prep_vst(load_lncrna(sp) |> select(feature_id, any_of(ids)), design)
  ids <- intersect(colnames(g$vst), colnames(l$vst))
  list(G = g$vst[, ids], L = l$vst[, ids], info = design_for(g$vst[, ids], design),
       n = tibble(species = sp, samples = length(ids), colonies = n_distinct(design_for(g$vst[, ids], design)$colony),
                  lncrna_in = l$n_raw, lncrna_kept = nrow(l$vst), genes_in = g$n_raw, genes_kept = nrow(g$vst),
                  pairs = nrow(l$vst) * nrow(g$vst)))
}
dat <- lapply(setNames(species_all, species_all), prep_layers)
n_tab <- bind_rows(lapply(dat, `[[`, "n"))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab, format.args = list(big.mark = ","))
```

## Model

```{r functions}
zrows <- function(m) { m <- m - rowMeans(m); m / sqrt(rowSums(m^2)) }   # unit-norm rows: r = crossprod

residualize_colony <- function(m, info) {
  X <- model.matrix(~ factor(info$colony))
  m - t(qr.fitted(qr(X), t(m)))
}

# Number of lncRNA x mRNA pairs with |r| >= each threshold. Rows of L and G are
# unit-norm, so r = L %*% t(G). Chunks keep memory near chunk x nrow(G) doubles.
count_edges <- function(L, G, chunk = 1000) {
  cnt <- numeric(length(thresholds))
  for (s in split(seq_len(nrow(L)), ceiling(seq_len(nrow(L)) / chunk))) {
    r <- abs(tcrossprod(L[s, , drop = FALSE], G))
    cnt <- cnt + vapply(thresholds, function(t) sum(r >= t), numeric(1))
  }
  setNames(cnt, paste0("r", thresholds))
}

perm_indices <- function(info, within_timepoint) {
  n <- nrow(info)
  lapply(seq_len(n_perm), function(b) {
    if (!within_timepoint) return(sample.int(n))
    idx <- seq_len(n)
    for (g in split(seq_len(n), info$timepoint)) idx[g] <- g[sample.int(length(g))]
    idx
  })
}

run_null <- function(L, G, perms) {
  res <- parallel::mclapply(perms, function(p) count_edges(L[, p, drop = FALSE], G),
                            mc.cores = n_workers, mc.preschedule = FALSE)
  bad <- vapply(res, function(x) is.null(x) || inherits(x, "try-error"), logical(1))
  if (any(bad)) stop(sum(bad), " permutation workers failed: ", paste(unique(unlist(lapply(res[bad], as.character))), collapse = " | "))
  do.call(rbind, res)
}

summarise_null <- function(obs, null_mat, sp, null_type, analysis) {
  q95 <- apply(null_mat, 2, quantile, 0.95)
  med <- apply(null_mat, 2, median)
  tibble(species = sp, analysis = analysis, null = null_type, threshold = thresholds,
         observed = obs, null_median = med, null_q95 = q95, null_max = apply(null_mat, 2, max),
         ratio_to_null_median = obs / pmax(1, med),
         p_emp = (1 + colSums(sweep(null_mat, 2, obs, `>=`))) / (1 + nrow(null_mat)),
         exceeds = obs > q95)
}

decide <- function(s) {
  v <- if (all(s$exceeds)) "supported" else if (any(!s$exceeds[s$threshold == 0.9])) "not supported" else "inconclusive"
  v
}
```

```{r run}
perms <- lapply(dat, function(d) list(full = perm_indices(d$info, FALSE), within_tp = perm_indices(d$info, TRUE)))

run_species <- function(sp, residualize) {
  d <- dat[[sp]]
  L <- d$L; G <- d$G
  if (residualize) { L <- residualize_colony(L, d$info); G <- residualize_colony(G, d$info) }
  L <- zrows(L); G <- zrows(G)
  analysis <- if (residualize) "colony-residualized" else "primary"
  obs <- count_edges(L, G)
  nf <- run_null(L, G, perms[[sp]]$full)
  nt <- run_null(L, G, perms[[sp]]$within_tp)
  list(summary = bind_rows(summarise_null(obs, nf, sp, "full permutation", analysis),
                           summarise_null(obs, nt, sp, "within-timepoint", analysis)),
       null = bind_rows(as_tibble(nf) |> mutate(perm = row_number(), null = "full permutation"),
                        as_tibble(nt) |> mutate(perm = row_number(), null = "within-timepoint")) |>
         mutate(species = sp, analysis = analysis, .before = 1))
}

runs <- c(lapply(species_all, run_species, residualize = FALSE),
          lapply(species_all, run_species, residualize = TRUE))
summ <- bind_rows(lapply(runs, `[[`, "summary"))
nulls <- bind_rows(lapply(runs, `[[`, "null"))
write_csv(summ, file.path(out_dir, "edge_counts_vs_null.csv"))
write_csv(nulls, file.path(out_dir, "null_counts.csv"))
```

## Result against the decision rule

```{r decide}
primary <- summ |> filter(analysis == "primary", null == "full permutation")
knitr::kable(primary, digits = 3, format.args = list(big.mark = ","))
verdicts <- summ |> group_by(analysis, null) |> group_modify(~ tibble(verdict = decide(.x))) |> ungroup()
write_csv(verdicts, file.path(out_dir, "verdicts.csv"))
```

**Primary verdict (full permutation): `r verdicts$verdict[verdicts$analysis == "primary" & verdicts$null == "full permutation"]`.**

```{r fig, fig.width = 10, fig.height = 6}
plot_null <- nulls |> pivot_longer(starts_with("r0"), names_to = "threshold", values_to = "count") |>
  mutate(threshold = as.numeric(sub("^r", "", threshold)))
g <- ggplot(plot_null, aes(factor(threshold), count + 1)) +
  geom_boxplot(aes(fill = null), outlier.size = 0.5, position = position_dodge(0.8)) +
  geom_point(data = summ |> distinct(species, analysis, threshold, observed),
             aes(factor(threshold), observed + 1), inherit.aes = FALSE, shape = 4, size = 3, stroke = 1.2) +
  scale_y_log10() + facet_grid(analysis ~ species) +
  scale_fill_manual(values = c(`full permutation` = "grey85", `within-timepoint` = "grey55")) +
  labs(x = "|r| threshold", y = "Pairs at or above threshold + 1 (log scale); x = observed", fill = NULL) +
  theme_bw()
ggsave(file.path(out_dir, "edges_vs_null.png"), g, width = 10, height = 6, dpi = 150)
g
```

## Sensitivity checks

Within-timepoint null and colony residualization, each with the same rule:

```{r sens}
knitr::kable(verdicts)
knitr::kable(summ |> filter(!(analysis == "primary" & null == "full permutation")) |>
               select(species, analysis, null, threshold, observed, null_median, null_q95, ratio_to_null_median, p_emp, exceeds),
             digits = 3, format.args = list(big.mark = ","))
```

## Caveats

- About 10 colonies per species, measured at four timepoints. The full
  permutation breaks colony and season structure that both layers share, so
  exceeding it shows shared structure, not necessarily lncRNA-specific
  co-regulation; the within-timepoint and colony-residualized nulls narrow
  that down.
- lncRNA features were already filtered upstream (fewer than 10 counts in
  more than 50% of samples removed); the H01 filter removes few more (see
  the n table).
- Edge counts at a threshold are not independent across pairs: one lncRNA
  correlated with a gene module contributes many edges.
- Species is confounded with reference quality (Peve N50 0.17 Mb; Ptuh mapped
  to *P. meandrina*), which affects lncRNA annotation.
