---
title: "H02: temporal plasticity across species"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2) })
design <- load_design()
orth <- load_orthologs()
out_dir <- file.path(root, "hypotheses", "H02", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
predicted <- c("Apul", "Ptuh", "Peve")          # predicted order, most to least plastic
alpha <- 0.05
set.seed(20261006)
```

## Pre-registered decision rule

> `supported` if the count of DE groups and the median max-|LFC| both follow
> Apul > Ptuh > Peve, and the pairwise Wilcoxon tests for adjacent species are
> significant at Holm-adjusted 0.05. `not supported` if either ordering is
> reversed between any pair with a significant test. Otherwise `inconclusive`.

Model: DESeq2 per species, `~ colony + timepoint`, LRT against `~ colony`, BH
FDR 0.05 within species. Secondary: per ortholog group, max |LFC| over the six
pairwise timepoint contrasts; Kruskal-Wallis across species, pairwise Wilcoxon
rank-sum with Holm correction. Multiple-testing correction: BH within species
for DE calls; Holm across the three species pairs.

Operational choices not spelled out in `hypothesis.md` (none changes the rule):

- **Common gene set.** The H01 filter is applied per species, then only groups
  passing in all three are kept, so DE counts and |LFC| distributions are
  compared over the same groups (the hypothesis states gene sets must be the
  same size). Per-species-filter counts are reported alongside.
- **"Adjacent species"** are Apul–Ptuh and Ptuh–Peve in the predicted order.
  "Reversed with a significant test" is checked for all three pairs: a pair
  whose Holm-adjusted Wilcoxon p < 0.05 and whose DE-count or median
  max-|LFC| ordering is opposite to the prediction.
- **LFCs** are the unshrunken DESeq2 MLEs from the full model.

## Data

```{r data}
counts_og <- lapply(setNames(species_all, species_all), function(sp) as_count_matrix(genes_by_ortholog(sp, orthologs = orth)))
prep <- lapply(counts_og, prep_counts, design = design)
common <- Reduce(intersect, lapply(prep, function(p) rownames(p$counts)))

n_tab <- bind_rows(lapply(species_all, function(sp) {
  p <- prep[[sp]]
  tibble(species = sp, samples = ncol(p$counts), colonies = n_distinct(p$info$colony),
         groups_in = p$n_raw, groups_pass_filter = nrow(p$counts), groups_common = length(common))
}))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab)
```

```{r design-table}
design_tab <- bind_rows(lapply(species_all, function(sp) prep[[sp]]$info)) |>
  count(species, timepoint) |>
  pivot_wider(names_from = timepoint, values_from = n, values_fill = 0)
write_csv(design_tab, file.path(out_dir, "design_per_timepoint.csv"))
knitr::kable(design_tab)
```

Library sizes (all genes, not only orthologs):

```{r libsize}
libsize <- bind_rows(lapply(species_all, function(sp) {
  g <- as_count_matrix(load_genes(sp))
  i <- design_for(g, design)
  g <- g[, i$has_genes %in% TRUE]
  tibble(species = sp, sample_id = colnames(g), libsize = colSums(g))
}))
write_csv(libsize, file.path(out_dir, "library_sizes.csv"))
knitr::kable(libsize |> group_by(species) |>
               summarise(min = min(libsize), median = median(libsize), max = max(libsize)),
             format.args = list(big.mark = ","))
```

## Model

```{r fit}
pairs_tp <- combn(c("TP1", "TP2", "TP3", "TP4"), 2, simplify = FALSE)

# Per species: LRT padj and max |LFC| over pairwise contrasts for the given groups.
run_species <- function(counts, info, groups) {
  dds <- fit_timepoint_deseq(counts[groups, , drop = FALSE], info)
  lrt <- DESeq2::results(dds, alpha = alpha)
  lfc <- sapply(pairs_tp, function(pr) DESeq2::results(dds, contrast = c("timepoint", pr[2], pr[1]))$log2FoldChange)
  colnames(lfc) <- vapply(pairs_tp, function(pr) paste0(pr[2], "_vs_", pr[1]), "")
  bind_cols(tibble(feature_id = rownames(lrt), baseMean = lrt$baseMean, lrt_stat = lrt$stat,
                   pvalue = lrt$pvalue, padj = lrt$padj),
            as_tibble(lfc)) |>
    mutate(max_abs_lfc = apply(abs(lfc), 1, max))
}

res <- bind_rows(lapply(species_all, function(sp)
  run_species(prep[[sp]]$counts, prep[[sp]]$info, common) |> mutate(species = sp, .before = 1)))
write_csv(res, file.path(out_dir, "de_per_group.csv"))

# Per-species-filter DE counts (each species on its own filtered set), descriptive only.
de_own <- bind_rows(lapply(species_all, function(sp) {
  r <- run_species(prep[[sp]]$counts, prep[[sp]]$info, rownames(prep[[sp]]$counts))
  tibble(species = sp, n_tested = nrow(r), n_de = sum(r$padj < alpha, na.rm = TRUE))
}))
write_csv(de_own, file.path(out_dir, "de_counts_own_filter.csv"))
```

```{r fig-lfc, fig.width = 7, fig.height = 4}
g <- ggplot(res, aes(factor(species, predicted), max_abs_lfc)) +
  geom_violin(scale = "width", fill = "grey85") +
  geom_boxplot(width = 0.12, outlier.shape = NA) +
  coord_cartesian(ylim = c(0, quantile(res$max_abs_lfc, 0.99, na.rm = TRUE))) +
  labs(x = NULL, y = "max |log2 FC| over pairwise timepoint contrasts") +
  theme_bw()
ggsave(file.path(out_dir, "max_abs_lfc_violin.png"), g, width = 7, height = 4, dpi = 150)
g
```

## Result against the decision rule

```{r decide}
summarise_plasticity <- function(r) {
  de <- r |> group_by(species) |>
    summarise(n_tested = sum(!is.na(padj)), n_de = sum(padj < alpha, na.rm = TRUE),
              median_max_abs_lfc = median(max_abs_lfc, na.rm = TRUE), .groups = "drop") |>
    rowwise() |>
    mutate(prop_de = n_de / n_tested,
           prop_lo = prop.test(n_de, n_tested)$conf.int[1],
           prop_hi = prop.test(n_de, n_tested)$conf.int[2]) |>
    ungroup()

  kw <- kruskal.test(max_abs_lfc ~ factor(species), data = r)

  pair_tab <- bind_rows(lapply(combn(predicted, 2, simplify = FALSE), function(pr) {
    x <- r$max_abs_lfc[r$species == pr[1]]; y <- r$max_abs_lfc[r$species == pr[2]]
    w <- wilcox.test(x, y, conf.int = TRUE)
    tibble(more = pr[1], less = pr[2], adjacent = abs(match(pr[1], predicted) - match(pr[2], predicted)) == 1,
           hl_shift = unname(w$estimate), hl_lo = w$conf.int[1], hl_hi = w$conf.int[2], p = w$p.value)
  })) |>
    mutate(p_holm = p.adjust(p, "holm"),
           count_order_ok = de$n_de[match(more, de$species)] > de$n_de[match(less, de$species)],
           lfc_order_ok = de$median_max_abs_lfc[match(more, de$species)] > de$median_max_abs_lfc[match(less, de$species)])

  ordered <- function(v) all(diff(v[match(predicted, de$species)]) < 0)
  verdict <-
    if (ordered(de$n_de) && ordered(de$median_max_abs_lfc) &&
        all(pair_tab$p_holm[pair_tab$adjacent] < 0.05)) "supported" else
    if (any(pair_tab$p_holm < 0.05 & !(pair_tab$count_order_ok & pair_tab$lfc_order_ok))) "not supported" else
    "inconclusive"
  list(de = de, kw = kw, pairs = pair_tab, verdict = verdict)
}

primary <- summarise_plasticity(res)
write_csv(primary$de, file.path(out_dir, "de_counts.csv"))
write_csv(primary$pairs, file.path(out_dir, "pairwise_wilcoxon.csv"))
knitr::kable(primary$de, digits = 3)
knitr::kable(primary$pairs, digits = 4)
```

Kruskal-Wallis across species: chi-squared = `r round(primary$kw$statistic, 1)`,
df = `r primary$kw$parameter`, p = `r signif(primary$kw$p.value, 3)`.
Wilcoxon effect sizes are Hodges-Lehmann shifts in max |log2 FC| (first species
minus second) with 95% CIs.

**Primary verdict: `r primary$verdict`.**

## Sensitivity checks

Each check re-applies the same decision rule. BH and Holm are recomputed within
the subset.

### Well-annotated orthologs

```{r sens-annot}
subset_rerun <- function(keep_groups) {
  r <- res |> filter(feature_id %in% keep_groups) |>
    group_by(species) |> mutate(padj = p.adjust(pvalue, "BH")) |> ungroup()
  summarise_plasticity(r)
}
ident_groups <- orth$group_id[!is.na(orth$avg_identity) & orth$avg_identity >= 60]
sp_groups <- orth$group_id[!is.na(orth$protein_name) & orth$protein_name != ""]

sens_ident <- subset_rerun(ident_groups)
sens_sp <- subset_rerun(sp_groups)
sens_tab <- bind_rows(
  sens_ident$de |> mutate(check = "avg_identity >= 60", verdict = sens_ident$verdict),
  sens_sp$de |> mutate(check = "SwissProt name", verdict = sens_sp$verdict))
write_csv(sens_tab, file.path(out_dir, "sens_annotation_de_counts.csv"))
write_csv(bind_rows(sens_ident$pairs |> mutate(check = "avg_identity >= 60"),
                    sens_sp$pairs |> mutate(check = "SwissProt name")),
          file.path(out_dir, "sens_annotation_pairwise_wilcoxon.csv"))
knitr::kable(sens_tab, digits = 3)
```

### Downsampled to equal depth

Each sample's counts are binomially thinned so that its full-transcriptome
library size equals the smallest per-species median
(`r format(min(tapply(libsize$libsize, libsize$species, median)), big.mark = ",")`).
Filtering and fitting are then repeated from scratch.

```{r sens-depth}
target <- min(tapply(libsize$libsize, libsize$species, median))
thin <- function(m, sp) {
  ls <- libsize$libsize[libsize$species == sp][match(colnames(m), libsize$sample_id[libsize$species == sp])]
  p <- pmin(1, target / ls)
  out <- m
  for (j in seq_len(ncol(m))) out[, j] <- rbinom(nrow(m), round(m[, j]), p[j])
  out
}
prep_ds <- lapply(species_all, function(sp) {
  p <- prep_counts(counts_og[[sp]], design)       # sample selection only; filter re-applied below
  m <- thin(round(counts_og[[sp]][, colnames(p$counts)]), sp)
  list(counts = filter_expressed(m), info = p$info)
}) |> setNames(species_all)
common_ds <- Reduce(intersect, lapply(prep_ds, function(p) rownames(p$counts)))
res_ds <- bind_rows(lapply(species_all, function(sp)
  run_species(prep_ds[[sp]]$counts, prep_ds[[sp]]$info, common_ds) |> mutate(species = sp, .before = 1)))
sens_ds <- summarise_plasticity(res_ds)
write_csv(sens_ds$de |> mutate(n_common = length(common_ds), verdict = sens_ds$verdict),
          file.path(out_dir, "sens_downsampled_de_counts.csv"))
write_csv(sens_ds$pairs, file.path(out_dir, "sens_downsampled_pairwise_wilcoxon.csv"))
knitr::kable(sens_ds$de, digits = 3)
```

```{r summary}
verdicts <- tibble(analysis = c("primary", "avg_identity >= 60", "SwissProt name", "downsampled"),
                   verdict = c(primary$verdict, sens_ident$verdict, sens_sp$verdict, sens_ds$verdict))
write_csv(verdicts, file.path(out_dir, "verdicts.csv"))
knitr::kable(verdicts)
```

## Caveats

- About 10 colonies per species; DE power differs with per-species n (40, 38,
  39 samples) and dispersion, so DE counts mix effect size and power.
- Max |LFC| over six contrasts uses unshrunken MLEs, which are inflated for
  low-count groups; species with lower depth (Peve) can look more variable
  for that reason alone. The downsampling check addresses depth, not
  shrinkage.
- The Wilcoxon tests treat ortholog groups as independent observations;
  they are not (co-regulated genes), so p-values are optimistic.
- Species is confounded with reference-genome quality (Peve N50 0.17 Mb; Ptuh
  mapped to *P. meandrina*); the annotation subsets are the pre-registered
  check.
