---
title: "H10: are ITS2 symbiont communities colony-fixed?"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2); library(vegan) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H10", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
n_perm <- 999
n_boot <- 500
set.seed(20261006)
```

## Pre-registered decision rule

> `supported` if R2(colony) > 0.5 and R2(timepoint) < 0.1 in all species, and
> within-colony distance is lower than between-colony with p < 0.01.
> `not supported` if R2(timepoint) >= R2(colony) in any species. Otherwise
> `inconclusive`. Secondary test reported but does not change the status.

Model: per species, `adonis2(bray ~ colony + timepoint, strata = colony)`,
999 permutations, on ITS2 profile relative abundances, all samples. Paired
Wilcoxon of within- vs between-colony distances. Secondary:
`lmer(trait ~ dominant_genus + timepoint + (1|colony))`.

Operational choices not spelled out in `hypothesis.md` (none changes the rule):

- **Samples.** All ITS2 samples, which cover many more colonies than the
  molecular layers (see the n table); colony, species and timepoint come from
  `config/design.csv`. Colonies sampled once contribute to PERMANOVA but not
  to the paired distance test.
- **PERMANOVA.** Terms are sequential (colony, then timepoint). R2 does not
  depend on permutations; the colony p-value uses unrestricted permutations,
  the timepoint p-value permutations blocked within colony. R2 gets a 95% CI
  from `r n_boot` bootstrap resamples of colonies (all samples of a drawn
  colony kept, duplicates relabelled as distinct colonies).
- **Paired distance test.** For each colony with >= 2 samples: within = mean
  Bray-Curtis between its own samples at different timepoints; between = mean
  distance from its samples to other colonies' samples at the same timepoint.
  One-sided Wilcoxon signed-rank (within < between) across colonies, per
  species; the rule is applied per species. Hodges-Lehmann shift with 95% CI.
- **Dominant genus** is the first letter of the sample's most abundant
  profile (A *Symbiodinium*, C *Cladocopium*, D *Durusdinium*, G
  *Gerakladium*). The secondary model is fit only where at least two genera
  each dominate >= 5 samples; traits `cells.cm2` and `calc.umol.cm2.hr`.

## Data

```{r data}
its2 <- load_its2()
prof_cols <- names(its2)[vapply(its2, is.numeric, logical(1))]
dat <- design |> filter(has_its2 %in% TRUE) |> select(sample_id, species, colony, timepoint) |>
  inner_join(its2 |> select(sample_id, all_of(prof_cols)), by = "sample_id")

by_sp <- lapply(setNames(species_all, species_all), function(sp) {
  d <- dat |> filter(species == sp)
  m <- as.matrix(d[, prof_cols]); rownames(m) <- d$sample_id
  m <- m[, colSums(m) > 0, drop = FALSE]
  dom <- colnames(m)[max.col(m, ties.method = "first")]
  list(info = d |> select(sample_id, species, colony, timepoint) |>
         mutate(dominant_profile = dom, dominant_genus = substr(dom, 1, 1)), m = m)
})

n_tab <- bind_rows(lapply(species_all, function(sp) {
  i <- by_sp[[sp]]$info
  top <- sort(table(i$dominant_profile), decreasing = TRUE)
  tibble(species = sp, samples = nrow(i), colonies = n_distinct(i$colony),
         colonies_ge2_samples = sum(table(i$colony) >= 2),
         profiles_present = ncol(by_sp[[sp]]$m), dominant_profiles = length(top),
         top_profile = names(top)[1], top_profile_share = as.numeric(top[1]) / nrow(i),
         genera = paste(sort(unique(i$dominant_genus)), collapse = ","))
}))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab, digits = 3)
```

Samples per timepoint:

```{r design-table}
tp_tab <- bind_rows(lapply(by_sp, `[[`, "info")) |> count(species, timepoint) |>
  pivot_wider(names_from = timepoint, values_from = n, values_fill = 0)
write_csv(tp_tab, file.path(out_dir, "samples_per_timepoint.csv"))
knitr::kable(tp_tab)
```

## Model

### PERMANOVA

```{r permanova}
r2_terms <- function(m, info) {
  a <- adonis2(vegdist(m, "bray") ~ colony + timepoint, data = info, by = "terms", permutations = 0)
  a$R2[1:2]
}

permanova_sp <- function(sp) {
  m <- by_sp[[sp]]$m
  info <- by_sp[[sp]]$info |> mutate(colony = factor(colony), timepoint = factor(timepoint)) |> as.data.frame()
  d <- vegdist(m, "bray")
  a_free <- adonis2(d ~ colony + timepoint, data = info, by = "terms", permutations = n_perm)
  a_block <- adonis2(d ~ colony + timepoint, data = info, by = "terms",
                     permutations = permute::how(blocks = info$colony, nperm = n_perm))
  cols <- unique(as.character(info$colony))
  boot <- replicate(n_boot, {
    draw <- sample(cols, replace = TRUE)
    idx <- unlist(lapply(seq_along(draw), function(k) which(info$colony == draw[k])))
    lab <- unlist(lapply(seq_along(draw), function(k) rep(paste0(draw[k], "_", k), sum(info$colony == draw[k]))))
    bi <- info[idx, ]; bi$colony <- factor(lab)
    r2_terms(m[idx, , drop = FALSE], bi)
  })
  tibble(species = sp, term = c("colony", "timepoint"), R2 = a_free$R2[1:2],
         R2_lo = apply(boot, 1, quantile, 0.025), R2_hi = apply(boot, 1, quantile, 0.975),
         p = c(a_free$`Pr(>F)`[1], a_block$`Pr(>F)`[2]), p_permutation = c("unrestricted", "within colony"))
}
perm <- bind_rows(lapply(species_all, permanova_sp))
write_csv(perm, file.path(out_dir, "permanova.csv"))
knitr::kable(perm, digits = 3)
```

### Within- vs between-colony distances

```{r paired}
paired_sp <- function(sp) {
  info <- by_sp[[sp]]$info
  D <- as.matrix(vegdist(by_sp[[sp]]$m, "bray"))
  per_col <- bind_rows(lapply(split(seq_len(nrow(info)), info$colony), function(i) {
    if (length(i) < 2) return(NULL)
    w <- D[i, i][upper.tri(D[i, i])]
    b <- unlist(lapply(i, function(s) {
      j <- which(info$timepoint == info$timepoint[s] & info$colony != info$colony[s])
      D[s, j]
    }))
    if (!length(b)) return(NULL)
    tibble(colony = info$colony[i[1]], n_samples = length(i), within = mean(w), between = mean(b))
  }))
  w <- wilcox.test(per_col$within, per_col$between, paired = TRUE, alternative = "less", conf.int = TRUE, exact = FALSE)
  w2 <- wilcox.test(per_col$within, per_col$between, paired = TRUE, conf.int = TRUE, exact = FALSE)
  list(per_colony = per_col |> mutate(species = sp, .before = 1),
       summary = tibble(species = sp, colonies = nrow(per_col),
                        median_within = median(per_col$within), median_between = median(per_col$between),
                        hl_shift = unname(w2$estimate), hl_lo = w2$conf.int[1], hl_hi = w2$conf.int[2],
                        p_one_sided = w$p.value))
}
pd <- lapply(species_all, paired_sp)
paired <- bind_rows(lapply(pd, `[[`, "summary"))
write_csv(paired, file.path(out_dir, "within_vs_between.csv"))
write_csv(bind_rows(lapply(pd, `[[`, "per_colony")), file.path(out_dir, "within_vs_between_per_colony.csv"))
knitr::kable(paired, digits = 4)
```

```{r fig-paired, fig.width = 9, fig.height = 3.6}
pc <- bind_rows(lapply(pd, `[[`, "per_colony")) |> pivot_longer(c(within, between), names_to = "comparison", values_to = "bray")
g <- ggplot(pc, aes(factor(comparison, c("within", "between")), bray, group = colony)) +
  geom_line(alpha = 0.3) + geom_point(alpha = 0.6, size = 1) + facet_wrap(~ species) +
  labs(x = NULL, y = "Mean Bray-Curtis dissimilarity") + theme_bw()
ggsave(file.path(out_dir, "within_vs_between.png"), g, width = 9, height = 3.6, dpi = 150)
g
```

## Result against the decision rule

```{r decide}
dec <- perm |> select(species, term, R2) |> pivot_wider(names_from = term, values_from = R2, names_prefix = "R2_") |>
  left_join(paired |> select(species, p_one_sided), by = "species") |>
  mutate(passes = R2_colony > 0.5 & R2_timepoint < 0.1 & p_one_sided < 0.01,
         reverses = R2_timepoint >= R2_colony)
verdict <- if (all(dec$passes)) "supported" else if (any(dec$reverses)) "not supported" else "inconclusive"
write_csv(dec, file.path(out_dir, "decision_per_species.csv"))
knitr::kable(dec, digits = 4)
```

**Verdict: `r verdict`.**

## Secondary: dominant genus and physiology

Reported only; does not change the status.

```{r secondary}
phys <- load_physiology() |> select(sample_id, cells.cm2, calc.umol.cm2.hr)
sec <- bind_rows(lapply(species_all, function(sp) {
  info <- by_sp[[sp]]$info |> left_join(phys, by = "sample_id")
  gen_n <- table(info$dominant_genus)
  if (sum(gen_n >= 5) < 2) return(tibble(species = sp, trait = NA_character_, note = paste("genera:", paste(names(gen_n), gen_n, sep = "=", collapse = ", "), "- not fitted")))
  keep <- names(gen_n)[gen_n >= 5]
  bind_rows(lapply(c("cells.cm2", "calc.umol.cm2.hr"), function(tr) {
    df <- info |> filter(dominant_genus %in% keep, !is.na(.data[[tr]])) |>
      mutate(y = .data[[tr]], dominant_genus = factor(dominant_genus), timepoint = factor(timepoint), colony = factor(colony))
    fit <- lmerTest::lmer(y ~ dominant_genus + timepoint + (1 | colony), data = df)
    co <- summary(fit)$coefficients
    rn <- grep("^dominant_genus", rownames(co), value = TRUE)
    ci <- confint(fit, parm = rn, method = "Wald")
    tibble(species = sp, trait = tr, n = nrow(df), contrast = sub("^dominant_genus", "", rn),
           reference = levels(df$dominant_genus)[1], estimate = co[rn, "Estimate"],
           lo = ci[, 1], hi = ci[, 2], p = co[rn, "Pr(>|t|)"],
           note = paste("genera:", paste(names(gen_n), gen_n, sep = "=", collapse = ", ")))
  }))
}))
write_csv(sec, file.path(out_dir, "secondary_genus_physiology.csv"))
knitr::kable(sec, digits = 4)
```

## Sensitivity checks

The pre-registered check is the profile count per species and how dominated
each species is by one profile (n table above): a species where nearly every
sample carries the same profile makes colony and timepoint R2 both small and
the test close to trivial.

## Caveats

- Colony counts here (`r paste(n_tab$species, n_tab$colonies, sep = " ", collapse = ", ")`)
  exceed the ~10 molecular colonies per species, but many colonies are not
  sampled at all four timepoints (see `samples_per_timepoint.csv`).
- SymPortal profiles summarize ITS2 type profiles; within-profile shifts in
  sequence abundance are invisible here.
- Relative abundances: a change in one profile forces changes in others.
- The colony bootstrap for R2 treats colonies as exchangeable within species,
  ignoring site.
