---
title: "H25: distal co-expression, lncRNA–gene vs gene–gene equivalence"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
source(file.path(root, "R", "neighbors.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H25", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
bins <- c(0, 2000, 10000, 50000, 100000)
bin_labels <- c("0-2 kb", "2-10 kb", "10-50 kb", "50-100 kb")
margin <- 0.25
B_colony <- 2000
B_scaffold <- 1000
min_span <- 500000
n_cores <- max(1, parallel::detectCores() - 1)
set.seed(25)
```

## Pre-registered decision rule

> `supported`: Holm-adjusted p_TOST < 0.05 (colony bootstrap) in every
> species. `not supported`: in any species, the 95% percentile colony-bootstrap
> interval for Δ lies wholly below −0.25 or wholly above +0.25. Otherwise
> `inconclusive` (failure to establish equivalence is not evidence for a
> transcript-class difference).

Δ = mean_b(L_b − G_b) / mean_b(G_b) over the bins (2, 10], (10, 50],
(50, 100] kb, where L_b is the median residual r of kept lncRNA–gene pairs
and G_b the weighted median of kept gene–gene orientations; species with
mean_b(G_b) < 0.01 are `inconclusive`. Everything below follows
`hypothesis.md` (completed 2026-10-09, `cb42e81`).

Implementation notes (not in `hypothesis.md`; none changes the rule):

- Quantile cut points use `quantile()` type 7; a value equal to a cut point
  goes to the lower stratum. Tertiles of gap are over all orientations in
  the bin (gene–gene pairs count twice), as in `feasibility_check.R`.
- Residual SD divides by n − rank of the colony + timepoint design.
- Weighted median: the smallest r at which cumulative weight reaches half
  the total.
- In a colony-bootstrap replicate the drawn colonies keep the full-data
  strata and weights; only r is recomputed. Residualization uses the drawn
  colony instances as levels.
- In a scaffold-block replicate, scaffolds are drawn from those holding at
  least one kept orientation.
- C1 and C2 reuse the same colony-bootstrap index draws as the primary.

## Data

```{r data}
cut_q <- function(x, ref, k) findInterval(x, quantile(ref, seq(0, 1, length.out = k + 1)[-c(1, k + 1)]), left.open = TRUE) + 1

prep_species <- function(sp) {
  ids <- design |> filter(species == sp, has_genes %in% TRUE, has_lncrna %in% TRUE) |> pull(sample_id)
  g <- prep_vst(load_genes(sp) |> select(feature_id, any_of(ids)), design)
  l <- prep_vst(load_lncrna(sp) |> select(feature_id, any_of(ids)), design)
  ids <- intersect(colnames(g$vst), colnames(l$vst))
  G <- g$vst[, ids]; L <- l$vst[, ids]
  info <- design_for(G, design) |> mutate(colony = factor(colony), timepoint = factor(timepoint))
  gc <- load_gene_coords(sp) |> filter(feature_id %in% rownames(G))
  lc <- lncrna_coords(sp) |> filter(feature_id %in% rownames(L))
  X <- model.matrix(~ colony + timepoint, data = info)
  rsd <- function(m) { r <- residualize(m, info, c("colony", "timepoint")); sqrt(rowSums(r^2) / (ncol(m) - qr(X)$rank)) }
  feat <- bind_rows(tibble(feature_id = rownames(G), class = "gene", mean = rowMeans(G), rsd = rsd(G)),
                    tibble(feature_id = rownames(L), class = "lncrna", mean = rowMeans(L), rsd = rsd(L)))
  span <- bind_rows(gc, lc) |> group_by(chrom) |> summarise(span = max(end), .groups = "drop")

  lg <- neighbor_pairs(lc, gc, max_dist = 100000, bins = bins, bin_labels = bin_labels) |>
    filter(!overlap, gap > 2000) |> mutate(class = "lncRNA-gene")
  gg <- neighbor_pairs(gc, gc, same = TRUE, max_dist = 100000, bins = bins, bin_labels = bin_labels) |>
    filter(!overlap, gap > 2000) |> mutate(class = "gene-gene")
  pairs <- bind_rows(lg, gg) |> mutate(pair = row_number(), bin = droplevels(bin)) |>
    left_join(span, by = "chrom")

  # orientations: lncRNA-gene once (anchor gene), gene-gene twice
  orient <- bind_rows(pairs |> filter(class == "lncRNA-gene") |> transmute(pair, class, bin, gap, chrom, span, anchor = b, partner = a),
                      pairs |> filter(class == "gene-gene") |> transmute(pair, class, bin, gap, chrom, span, anchor = a, partner = b),
                      pairs |> filter(class == "gene-gene") |> transmute(pair, class, bin, gap, chrom, span, anchor = b, partner = a))
  fm <- setNames(feat$mean, feat$feature_id); fr <- setNames(feat$rsd, feat$feature_id)
  gmean <- feat$mean[feat$class == "gene"]
  orient <- orient |> group_by(bin) |> mutate(gap_t = cut_q(gap, gap, 3)) |> ungroup() |>
    mutate(anchor_t = cut_q(fm[anchor], gmean, 3))

  assign_strata <- function(o, k) {
    o |> mutate(pm = cut_q(fm[partner], feat$mean, k), pr = cut_q(fr[partner], feat$rsd, k),
                stratum = paste(bin, gap_t, anchor_t, pm, pr, sep = "|")) |>
      group_by(stratum) |> mutate(n_lg = sum(class == "lncRNA-gene"), n_gg = sum(class == "gene-gene")) |> ungroup() |>
      mutate(kept = n_lg >= 3 & n_gg >= 3)
  }
  o5 <- assign_strata(orient, 5); o3 <- assign_strata(orient, 3)
  share5 <- o5 |> filter(class == "lncRNA-gene") |> group_by(bin) |> summarise(share = mean(kept), .groups = "drop")
  level <- setNames(ifelse(share5$share < 0.5, "tertile", "quintile"), share5$bin)
  orient <- bind_rows(lapply(names(level), function(b) (if (level[[b]] == "quintile") o5 else o3) |> filter(bin == b))) |>
    filter(kept) |>
    mutate(w = ifelse(class == "lncRNA-gene", 1, n_lg / n_gg), bin = factor(bin, levels = levels(pairs$bin)))

  match_tab <- orient |> group_by(bin) |>
    summarise(strata_kept = n_distinct(stratum), lncrna_gene_kept = sum(class == "lncRNA-gene"),
              gene_gene_orientations_kept = sum(class == "gene-gene"), .groups = "drop") |>
    left_join(pairs |> filter(class == "lncRNA-gene") |> count(bin, name = "lncrna_gene_all"), by = "bin") |>
    mutate(species = sp, .before = 1, level = level[as.character(bin)], share_kept = lncrna_gene_kept / lncrna_gene_all)

  list(species = sp, G = G, L = L, info = info, pairs = pairs, orient = orient, match = match_tab,
       n = tibble(species = sp, samples = length(ids), colonies = n_distinct(info$colony),
                  sites = n_distinct(info$site), genes = nrow(G), lncrna = nrow(L),
                  lncrna_gene_pairs = sum(pairs$class == "lncRNA-gene"), gene_gene_pairs = sum(pairs$class == "gene-gene"),
                  scaffolds = n_distinct(pairs$chrom)))
}

dat <- lapply(setNames(species_all, species_all), prep_species)
n_tab <- bind_rows(lapply(dat, `[[`, "n"))
match_tab <- bind_rows(lapply(dat, `[[`, "match"))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
write_csv(match_tab, file.path(out_dir, "matching_per_bin.csv"))
knitr::kable(n_tab, format.args = list(big.mark = ","))
knitr::kable(match_tab, digits = 3, format.args = list(big.mark = ","))
```

## Model

```{r functions}
wmedian <- function(x, w) { o <- order(x); x <- x[o]; cw <- cumsum(w[o]); x[which(cw >= cw[length(cw)] / 2)[1]] }

# r for every pair given a sample index vector `cols` (columns, possibly
# repeated) and colony labels for those columns.
pair_correlations <- function(d, cols, colony_lab, terms) {
  info <- d$info[cols, ] |> mutate(colony = factor(colony_lab))
  A <- zrows(residualize(d$L[, cols, drop = FALSE], info, terms))
  Gz <- zrows(residualize(d$G[, cols, drop = FALSE], info, terms))
  p <- d$pairs
  r <- numeric(nrow(p))
  lg <- p$class == "lncRNA-gene"
  r[lg] <- pair_r(A, Gz, match(p$a[lg], rownames(A)), match(p$b[lg], rownames(Gz)))
  r[!lg] <- pair_r(Gz, Gz, match(p$a[!lg], rownames(Gz)), match(p$b[!lg], rownames(Gz)))
  r
}

# Matched per-bin medians and delta for a set of orientations `o` (rows of
# d$orient, possibly repeated) and pair correlations `r`.
delta_stat <- function(o, r) {
  per_bin <- o |> mutate(r = r[pair]) |> group_by(bin) |>
    summarise(L = median(r[class == "lncRNA-gene"]),
              G = wmedian(r[class == "gene-gene"], w[class == "gene-gene"]), .groups = "drop") |>
    mutate(diff = L - G)
  list(per_bin = per_bin, delta = mean(per_bin$diff) / mean(per_bin$G), G_mean = mean(per_bin$G))
}

boot_colony_idx <- function(d) {
  cl <- d$info |> distinct(colony, site)
  replicate(B_colony, {
    drawn <- unlist(lapply(split(as.character(cl$colony), cl$site), function(x) sample(x, length(x), replace = TRUE)))
    cols <- unlist(lapply(seq_along(drawn), function(k) which(d$info$colony == drawn[k])))
    lab <- unlist(lapply(seq_along(drawn), function(k) rep(k, sum(d$info$colony == drawn[k]))))
    list(cols = cols, lab = lab)
  }, simplify = FALSE)
}

tost <- function(boot, est) {
  p_lower <- (1 + sum(boot <= -margin)) / (length(boot) + 1)
  p_upper <- (1 + sum(boot >= margin)) / (length(boot) + 1)
  tibble(delta = est, delta_lo = quantile(boot, 0.025), delta_hi = quantile(boot, 0.975),
         p_lower = p_lower, p_upper = p_upper, p_tost = max(p_lower, p_upper))
}

# Run the primary (colony + timepoint) and checks C1 (colony only) and C2
# (scaffold span >= 500 kb) on the same colony-bootstrap draws.
analyse_species <- function(d) {
  full <- seq_len(nrow(d$info))
  r_ct <- pair_correlations(d, full, as.character(d$info$colony), c("colony", "timepoint"))
  r_c <- pair_correlations(d, full, as.character(d$info$colony), "colony")
  o_all <- d$orient; o_long <- d$orient |> filter(span >= min_span)
  sets <- list(primary = list(o = o_all, r = r_ct, terms = c("colony", "timepoint")),
               C1_colony_only = list(o = o_all, r = r_c, terms = "colony"),
               C2_span_500kb = list(o = o_long, r = r_ct, terms = c("colony", "timepoint")))
  obs <- lapply(sets, function(s) delta_stat(s$o, s$r))
  idx <- boot_colony_idx(d)
  boot <- parallel::mclapply(idx, function(b) {
    rr <- list(ct = pair_correlations(d, b$cols, b$lab, c("colony", "timepoint")),
               c = pair_correlations(d, b$cols, b$lab, "colony"))
    lapply(names(sets), function(nm) {
      s <- delta_stat(sets[[nm]]$o, if (nm == "C1_colony_only") rr$c else rr$ct)
      c(delta = s$delta, setNames(s$per_bin$diff, paste0("diff_", s$per_bin$bin)),
        setNames(s$per_bin$L, paste0("L_", s$per_bin$bin)), setNames(s$per_bin$G, paste0("G_", s$per_bin$bin)))
    }) |> setNames(names(sets))
  }, mc.cores = n_cores)
  if (any(vapply(boot, inherits, TRUE, "try-error"))) stop("colony bootstrap worker failed")

  # C3: scaffold-block bootstrap, primary r fixed
  by_scaf <- split(seq_len(nrow(o_all)), o_all$chrom)
  scafs <- names(by_scaf)
  scaf_draws <- replicate(B_scaffold, sample(scafs, length(scafs), replace = TRUE), simplify = FALSE)
  scaf_boot <- vapply(scaf_draws, function(sc) delta_stat(o_all[unlist(by_scaf[sc]), ], r_ct)$delta, 0)

  summ <- bind_rows(lapply(names(sets), function(nm) {
    bd <- vapply(boot, function(b) b[[nm]][["delta"]], 0)
    tost(bd, obs[[nm]]$delta) |> mutate(analysis = nm, G_mean = obs[[nm]]$G_mean,
                                        orientations = nrow(sets[[nm]]$o), .before = 1)
  }), tost(scaf_boot, obs$primary$delta) |> mutate(analysis = "C3_scaffold_block", G_mean = obs$primary$G_mean,
                                                   orientations = nrow(o_all), .before = 1)) |>
    mutate(species = d$species, .before = 1)

  per_bin <- bind_rows(lapply(names(sets), function(nm) {
    m <- do.call(rbind, lapply(boot, function(b) b[[nm]]))
    obs[[nm]]$per_bin |> rowwise() |>
      mutate(L_lo = quantile(m[, paste0("L_", bin)], 0.025), L_hi = quantile(m[, paste0("L_", bin)], 0.975),
             G_lo = quantile(m[, paste0("G_", bin)], 0.025), G_hi = quantile(m[, paste0("G_", bin)], 0.975),
             diff_lo = quantile(m[, paste0("diff_", bin)], 0.025), diff_hi = quantile(m[, paste0("diff_", bin)], 0.975)) |>
      ungroup() |> mutate(analysis = nm, .before = 1)
  })) |> mutate(species = d$species, .before = 1)

  draws <- tibble(species = d$species, replicate = seq_len(B_colony),
                  primary = vapply(boot, function(b) b$primary[["delta"]], 0),
                  C1_colony_only = vapply(boot, function(b) b$C1_colony_only[["delta"]], 0),
                  C2_span_500kb = vapply(boot, function(b) b$C2_span_500kb[["delta"]], 0))
  list(summary = summ, per_bin = per_bin, draws = draws, scaf_draws = tibble(species = d$species, delta = scaf_boot))
}
```

```{r run}
res <- lapply(dat, analyse_species)
summ <- bind_rows(lapply(res, `[[`, "summary"))
per_bin <- bind_rows(lapply(res, `[[`, "per_bin"))
write_csv(per_bin, file.path(out_dir, "per_bin_C4.csv"))
write_csv(bind_rows(lapply(res, `[[`, "draws")), file.path(out_dir, "colony_bootstrap_draws.csv"))
write_csv(bind_rows(lapply(res, `[[`, "scaf_draws")), file.path(out_dir, "scaffold_bootstrap_draws.csv"))

decide <- function(s) {
  s <- s |> mutate(no_reference = G_mean < 0.01,
                   p_tost_holm = p.adjust(p_tost, "holm"),
                   equivalent = !no_reference & p_tost_holm < 0.05,
                   beyond_margin = !no_reference & (delta_hi < -margin | delta_lo > margin))
  verdict <- if (all(s$equivalent)) "supported" else if (any(s$beyond_margin)) "not supported" else "inconclusive"
  list(per_species = s, verdict = verdict)
}
decisions <- lapply(split(summ, summ$analysis), decide)
summ_dec <- bind_rows(lapply(decisions, `[[`, "per_species"))
write_csv(summ_dec, file.path(out_dir, "delta_summary.csv"))
primary <- decisions$primary
write_csv(primary$per_species, file.path(out_dir, "decision_per_species.csv"))
knitr::kable(primary$per_species |> select(species, G_mean, delta, delta_lo, delta_hi, p_lower, p_upper, p_tost, p_tost_holm, equivalent, beyond_margin), digits = 4)
```

**Primary verdict: `r primary$verdict`.**

```{r fig, fig.width = 10, fig.height = 4}
fig <- per_bin |> filter(analysis == "primary") |>
  select(species, bin, L, L_lo, L_hi, G, G_lo, G_hi) |>
  pivot_longer(c(L, G), names_to = "class", values_to = "median_r") |>
  mutate(lo = ifelse(class == "L", L_lo, G_lo), hi = ifelse(class == "L", L_hi, G_hi),
         class = ifelse(class == "L", "lncRNA–gene", "gene–gene (matched)")) |>
  select(species, bin, class, median_r, lo, hi)
write_csv(fig, file.path(out_dir, "per_bin_figure_data.csv"))
g <- ggplot(fig, aes(bin, median_r, colour = class, group = class)) +
  geom_pointrange(aes(ymin = lo, ymax = hi), position = position_dodge(0.4)) +
  geom_hline(yintercept = 0, colour = "grey70") + facet_wrap(~ species) +
  scale_colour_manual(values = c("lncRNA–gene" = "black", "gene–gene (matched)" = "grey55")) +
  labs(x = "Gap between pair members", y = "Median residual r (95% colony bootstrap)", colour = NULL) +
  theme_bw() + theme(legend.position = "bottom")
ggsave(file.path(out_dir, "per_bin_medians.png"), g, width = 10, height = 4, dpi = 150)
g
```

## Sensitivity checks

C1 (colony-only residualization), C2 (scaffolds spanning >= 500 kb) and C3
(scaffold-block bootstrap in place of the colony bootstrap) are descriptive.

```{r checks}
knitr::kable(summ_dec |> filter(analysis != "primary") |>
               select(analysis, species, orientations, G_mean, delta, delta_lo, delta_hi, p_tost, p_tost_holm, equivalent, beyond_margin), digits = 4)
verdicts <- tibble(analysis = names(decisions), verdict = vapply(decisions, `[[`, "", "verdict")) |>
  arrange(match(analysis, c("primary", "C1_colony_only", "C2_span_500kb", "C3_scaffold_block")))
write_csv(verdicts, file.path(out_dir, "verdicts.csv"))
knitr::kable(verdicts)
```

C4, per-bin medians and differences with colony-bootstrap intervals:

```{r c4}
knitr::kable(per_bin |> filter(analysis == "primary") |> select(species, bin, L, G, diff, diff_lo, diff_hi), digits = 4)
```

## Caveats

- About 10 colonies per species; the colony bootstrap resamples 5 colonies
  per site.
- lncRNA strand is unknown (D-014) and gene models are incomplete.
- Matching keeps the lncRNAs whose expression overlaps genes'; see
  `matching_per_bin.csv` for the share kept.
- Pair correlations share members; the two bootstraps address sample-level
  and genomic dependence separately.
- Follow-up on the same data as H17, not an independent replication.
- Species is confounded with reference quality (Peve N50 0.17 Mb; Ptuh mapped
  to *P. meandrina*).
