---
title: "H01: colony vs season variance in gene expression"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
source(file.path(root, "R", "varpart.R"))
suppressPackageStartupMessages({
  library(tidyr); library(ggplot2); library(variancePartition); library(BiocParallel)
})
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H01", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
n_boot <- 1000
set.seed(20261006)
bp <- MulticoreParam(workers = max(1, parallel::detectCores() - 1), progressbar = FALSE)
```

## Pre-registered decision rule

> `supported` if median(colony fraction) > median(timepoint fraction) with
> non-overlapping bootstrap 95% CIs in all three species. `not supported` if
> timepoint >= colony in any species with non-overlapping CIs. Otherwise
> `inconclusive`.

Model: `~ (1|colony) + (1|timepoint) + (1|site)` per gene, per species, on
DESeq2 `vst` of genes with >= 10 counts in >= 25% of samples. 95% intervals on
the median fractions from 1000 bootstrap resamples of genes. No hypothesis
tests are run, so no multiple-testing correction applies.

## Data

```{r data}
prep <- lapply(setNames(species_all, species_all), function(sp) prep_vst(load_genes(sp), design))

n_tab <- bind_rows(lapply(species_all, function(sp) {
  p <- prep[[sp]]
  tibble(species = sp, samples = ncol(p$vst), colonies = n_distinct(p$info$colony),
         genes_in = p$n_raw, genes_kept = nrow(p$vst))
}))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab)
```

Design per species (samples per colony x timepoint; a 0 is a missing sample):

```{r design-table}
design_tab <- bind_rows(lapply(species_all, function(sp) prep[[sp]]$info)) |>
  count(species, site, colony, timepoint) |>
  pivot_wider(names_from = timepoint, values_from = n, values_fill = 0) |>
  arrange(species, site, colony)
write_csv(design_tab, file.path(out_dir, "design_per_colony.csv"))
knitr::kable(design_tab)
```

## Model

```{r fit}
form_full <- ~ (1 | colony) + (1 | timepoint) + (1 | site)

fit_vp <- function(v, info, form) fit_varpart(v, info, form, BPPARAM = bp)

vp_full <- bind_rows(lapply(species_all, function(sp)
  fit_vp(prep[[sp]]$vst, prep[[sp]]$info, form_full) |> mutate(species = sp, .before = 1)))
write_csv(vp_full, file.path(out_dir, "varpart_per_gene.csv"))
```

```{r fig-varpart, fig.width = 9, fig.height = 4}
long <- vp_full |>
  pivot_longer(c(colony, timepoint, site, Residuals), names_to = "component", values_to = "fraction") |>
  mutate(component = factor(component, c("colony", "timepoint", "site", "Residuals")))
g <- ggplot(long, aes(component, fraction)) +
  geom_violin(scale = "width", fill = "grey85") +
  geom_boxplot(width = 0.12, outlier.shape = NA) +
  facet_wrap(~ species) +
  labs(x = NULL, y = "Fraction of expression variance") +
  theme_bw()
ggsave(file.path(out_dir, "varpart_violin.png"), g, width = 9, height = 4, dpi = 150)
write_csv(long |> group_by(species, component) |>
            summarise(median = median(fraction), q25 = quantile(fraction, 0.25),
                      q75 = quantile(fraction, 0.75), .groups = "drop"),
          file.path(out_dir, "varpart_violin_summary.csv"))
g
```

## Result against the decision rule

```{r bootstrap}
boot_medians <- function(vp) boot_median_ci(vp, B = n_boot)

decide <- function(ci) {
  per_sp <- ci |>
    select(species, component, median, lo, hi) |>
    pivot_wider(names_from = component, values_from = c(median, lo, hi)) |>
    mutate(colony_wins = median_colony > median_timepoint & lo_colony > hi_timepoint,
           timepoint_wins = median_timepoint >= median_colony & lo_timepoint > hi_colony)
  verdict <- if (all(per_sp$colony_wins)) "supported" else
    if (any(per_sp$timepoint_wins)) "not supported" else "inconclusive"
  list(per_species = per_sp, verdict = verdict)
}

ci_full <- boot_medians(vp_full)
write_csv(ci_full, file.path(out_dir, "median_fraction_ci.csv"))
dec_full <- decide(ci_full)
write_csv(dec_full$per_species, file.path(out_dir, "decision_per_species.csv"))
knitr::kable(dec_full$per_species, digits = 3)
```

**Primary verdict: `r dec_full$verdict`.**

## Sensitivity checks

### Site dropped

Site is colony-level, so variance it absorbs would otherwise sit with colony.

```{r sens-nosite}
vp_nosite <- bind_rows(lapply(species_all, function(sp)
  fit_vp(prep[[sp]]$vst, prep[[sp]]$info, ~ (1 | colony) + (1 | timepoint)) |>
    mutate(species = sp, .before = 1)))
ci_nosite <- boot_medians(vp_nosite)
dec_nosite <- decide(ci_nosite)
write_csv(ci_nosite, file.path(out_dir, "sens_nosite_median_fraction_ci.csv"))
knitr::kable(dec_nosite$per_species, digits = 3)
```

Verdict without site: `r dec_nosite$verdict`.

### Three-way orthologs only

Same filter and model on the shared ortholog set, to check that the result is
not driven by species differences in reference-genome quality.

```{r sens-ortho}
orth <- load_orthologs()
vp_ortho <- bind_rows(lapply(species_all, function(sp) {
  p <- prep_vst(genes_by_ortholog(sp, orthologs = orth), design)
  fit_vp(p$vst, p$info, form_full) |> mutate(species = sp, .before = 1, n_genes = nrow(p$vst))
}))
ci_ortho <- boot_medians(vp_ortho)
dec_ortho <- decide(ci_ortho)
write_csv(ci_ortho, file.path(out_dir, "sens_ortho_median_fraction_ci.csv"))
knitr::kable(dec_ortho$per_species, digits = 3)
```

Verdict on orthologs: `r dec_ortho$verdict`.

```{r summary}
write_csv(tibble(analysis = c("primary", "site dropped", "three-way orthologs"),
                 verdict = c(dec_full$verdict, dec_nosite$verdict, dec_ortho$verdict)),
          file.path(out_dir, "verdicts.csv"))
```

## Caveats

- About 10 colonies per species; colony and timepoint fractions are each
  estimated from few levels (10 and 4), so per-gene estimates are noisy and
  the bootstrap over genes does not capture colony-sampling uncertainty.
- Timepoint is fit as a random effect with only 4 levels, as pre-registered.
- Site has 2 levels with 5 colonies each and is nested in colony.
- Missing samples (Peve TP3/TP4, Ptuh TP4) leave the design slightly unbalanced; see the design table.
