---
title: "H18: methylation and cis-lncRNA channels on within-colony expression"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
source(file.path(root, "R", "methylation.R"))
source(file.path(root, "R", "neighbors.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2); library(lme4) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H18", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
tps <- c("TP1", "TP2", "TP3", "TP4")
n_perm <- 1000
n_boot <- 2000
n_cores <- max(1, parallel::detectCores() - 1)
set.seed(20261007)
```

## Pre-registered decision rule

Amendment 1 (`hypothesis.md`, written before any analysis) makes the rule
operational; per species, in this order:

> 1. *No signal*: unique_M and unique_L both indistinguishable from null.
> 2. *Hierarchical*: path a (M_L → L) colony-bootstrap CI excludes 0, the
>    indirect effect a×b CI excludes 0, and a×b / c >= 0.25 where c is the
>    total effect of M_L on E (same sign as a×b).
> 3. *Independent*: both unique fractions > 0 and shared < 25% of the
>    combined-model R².
> 4. *Redundant*: shared >= the larger unique fraction and the combined R²
>    exceeds the 95th percentile of the combined R² under each of the two
>    nulls.
> 5. Otherwise *unclassified*.
>
> `supported` if >= 2 species are independent. `not supported` if >= 2 are
> redundant, or >= 2 are hierarchical. `inconclusive` if >= 2 have no
> signal, or otherwise.

"> 0" means above the permutation null, one-sided p < 0.05. E: gene vst;
M: gene-body methylation; L: vst of the nearest non-overlapping lncRNA within
10 kb; M_L: methylation over that lncRNA's locus. Every variable is centered
on its gene×colony mean and standardized within gene.

Implementation notes (not in `hypothesis.md`; none changes the rule):

- **Mixed model vs OLS.** Every variable is centered on its own
  gene×colony mean over the same samples, so the outcome has zero mean in
  every gene and colony and the `(1|colony) + (1|gene)` intercepts are
  structurally zero. The pre-registered `lmer` fits and their Nakagawa
  marginal R² are computed on the observed data and compared with OLS below.
  If they agree to within 1e-4 for every R² and commonality fraction, the
  permutation and bootstrap refits use the closed-form OLS solution (from
  the correlation matrix of E, M, L, M_L). If not, the rule is applied to
  the `lmer` values and RESULT.md says so. (H16 needed `lmer` throughout
  because its lagged outcome was not centered on the same samples; that
  does not apply here.)
- Pairs: for each gene with a GBM value, the nearest lncRNA (smallest gap,
  gap > 0) within 10 kb, using `neighbor_pairs()` as in H17. Ties are
  broken by lncRNA id.
- Permutation pairings exclude the identity; a colony contributes the
  timepoints it shares with its partner.
- Bootstrap: colonies resampled with replacement; 95% percentile intervals.

## Data

```{r data}
prep_species <- function(sp, min_cpg = 5) {
  cpg <- as_count_matrix(load_cpg(sp))
  ids <- design |> filter(species == sp, has_cpg %in% TRUE, has_lncrna %in% TRUE, has_genes %in% TRUE,
                          sample_id %in% colnames(cpg)) |> pull(sample_id)
  g <- prep_vst(load_genes(sp) |> select(feature_id, any_of(ids)), design)
  l <- prep_vst(load_lncrna(sp) |> select(feature_id, any_of(ids)), design)
  ids <- Reduce(intersect, list(ids, colnames(g$vst), colnames(l$vst)))
  info <- design |> filter(sample_id %in% ids) |> arrange(colony, timepoint)
  ids <- info$sample_id
  gcoord <- load_gene_coords(sp); lcoord <- lncrna_coords(sp)
  gb <- gene_body_methylation(cpg[, ids, drop = FALSE], gcoord, min_cpg = min_cpg)
  lb <- gene_body_methylation(cpg[, ids, drop = FALSE], lcoord, min_cpg = min_cpg)
  genes_gbm <- intersect(rownames(gb$gbm), rownames(g$vst))
  pr <- neighbor_pairs(gcoord |> filter(feature_id %in% genes_gbm), lcoord |> filter(feature_id %in% rownames(l$vst))) |>
    filter(!overlap, gap <= 10000) |> arrange(a, gap, b) |> distinct(a, .keep_all = TRUE) |>
    transmute(gene = a, lncrna = b, gap, has_ml = lncrna %in% rownames(lb$gbm), n_cpg = gb$n_cpg[gene])
  cols <- unique(info$colony)
  arr <- function(m, rows) setNames(lapply(cols, function(c) {
    out <- matrix(NA_real_, length(rows), 4, dimnames = list(NULL, tps))
    s <- info |> filter(colony == c)
    ok <- rows %in% rownames(m)
    out[ok, s$timepoint] <- m[rows[ok], s$sample_id, drop = FALSE]
    out
  }), cols)
  # included vs excluded GBM genes (pre-registered description)
  incl <- tibble(gene = genes_gbm, included = genes_gbm %in% pr$gene,
                 gbm = rowMeans(gb$gbm[genes_gbm, , drop = FALSE]), expr = rowMeans(g$vst[genes_gbm, ids, drop = FALSE])) |>
    group_by(included) |> summarise(genes = n(), mean_gbm = mean(gbm), median_gbm = median(gbm),
                                    mean_vst = mean(expr), median_vst = median(expr), .groups = "drop") |>
    mutate(species = sp, .before = 1)
  list(species = sp, cols = cols, pairs = pr,
       E = arr(g$vst, pr$gene), L = arr(l$vst, pr$lncrna), M = arr(gb$gbm, pr$gene), ML = arr(lb$gbm, pr$lncrna),
       incl = incl,
       n = tibble(species = sp, samples = length(ids), colonies = length(cols), genes_with_gbm = length(genes_gbm),
                  pairs = nrow(pr), pairs_with_lncrna_methylation = sum(pr$has_ml)))
}

dat <- lapply(setNames(species_all, species_all), prep_species)
n_tab <- bind_rows(lapply(dat, `[[`, "n"))
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab, format.args = list(big.mark = ","))
```

GBM genes with and without a cis-lncRNA partner (pre-registered check of the
selected subset):

```{r included}
incl_tab <- bind_rows(lapply(dat, `[[`, "incl"))
write_csv(incl_tab, file.path(out_dir, "included_vs_excluded_genes.csv"))
knitr::kable(incl_tab, digits = 3, format.args = list(big.mark = ","))
```

## Model

```{r functions}
# Long vectors for a pairing: unit i takes M and ML from colony mc[i], L from
# lc[i], E from ec[i], on the timepoints all three have. Center per pair x
# unit, then standardize per pair across units. `rows` subsets pairs.
make_long <- function(d, mc, lc, ec, rows = seq_len(nrow(d$pairs))) {
  center <- function(m) m - rowMeans(m, na.rm = TRUE)
  parts <- lapply(seq_along(mc), function(i) {
    E <- d$E[[ec[i]]][rows, , drop = FALSE]; L <- d$L[[lc[i]]][rows, , drop = FALSE]
    M <- d$M[[mc[i]]][rows, , drop = FALSE]; ML <- d$ML[[mc[i]]][rows, , drop = FALSE]
    ok <- !is.na(E[1, ]) & !is.na(L[1, ]) & !is.na(M[1, ])
    if (sum(ok) < 2) return(NULL)
    list(E = center(E[, ok, drop = FALSE]), L = center(L[, ok, drop = FALSE]),
         M = center(M[, ok, drop = FALSE]), ML = center(ML[, ok, drop = FALSE]), unit = i)
  })
  parts <- Filter(Negate(is.null), parts)
  bind <- function(v) do.call(cbind, lapply(parts, `[[`, v))
  E <- bind("E"); L <- bind("L"); M <- bind("M"); ML <- bind("ML")
  std <- function(m) { s <- apply(m, 1, sd, na.rm = TRUE); s[!is.finite(s) | s == 0] <- NA; m / s }
  E <- std(E); L <- std(L); M <- std(M); ML <- std(ML)
  unit <- rep(sapply(parts, `[[`, "unit"), times = sapply(parts, function(p) ncol(p$E)))
  tibble(pair = rep(rows, times = ncol(E)), unit = rep(unit, each = nrow(E)),
         E = as.vector(E), L = as.vector(L), M = as.vector(M), ML = as.vector(ML))
}

r2_from <- function(rEM, rEL, rML) (rEM^2 + rEL^2 - 2 * rEM * rEL * rML) / (1 - rML^2)

# Commonality and path statistics from long data (closed-form OLS).
stats_of <- function(x) {
  x1 <- x |> filter(is.finite(E), is.finite(L), is.finite(M))
  cm <- cor(x1[, c("E", "M", "L")])
  R2_ML <- r2_from(cm["E", "M"], cm["E", "L"], cm["M", "L"]); R2_M <- cm["E", "M"]^2; R2_L <- cm["E", "L"]^2
  x2 <- x |> filter(is.finite(E), is.finite(L), is.finite(ML))
  cp <- cor(x2[, c("E", "L", "ML")])
  a <- cp["L", "ML"]; b <- (cp["E", "L"] - cp["E", "ML"] * a) / (1 - a^2); c <- cp["E", "ML"]
  c(R2_combined = R2_ML, R2_M = R2_M, R2_L = R2_L,
    unique_M = R2_ML - R2_L, unique_L = R2_ML - R2_M, shared = R2_M + R2_L - R2_ML,
    path_a = a, path_b = b, indirect = a * b, total_c = c, prop_mediated = a * b / c,
    rows = nrow(x1), rows_path = nrow(x2))
}

# Pre-registered lmer fits on the observed data (for the OLS equivalence check).
lmer_check <- function(x, d) {
  x <- x |> filter(is.finite(E), is.finite(L), is.finite(M)) |>
    mutate(colony = d$cols[unit], gene = factor(pair))
  r2m <- function(f) {
    fit <- lmer(f, data = x, control = lmerControl(calc.derivs = FALSE))
    vf <- var(as.vector(model.matrix(fit) %*% fixef(fit)))
    vc <- as.data.frame(VarCorr(fit))$vcov
    c(R2 = vf / (vf + sum(vc)), singular = isSingular(fit))
  }
  a <- r2m(E ~ M + L + (1 | colony) + (1 | gene)); m <- r2m(E ~ M + (1 | colony) + (1 | gene)); l <- r2m(E ~ L + (1 | colony) + (1 | gene))
  tibble(R2_combined_lmer = a[["R2"]], R2_M_lmer = m[["R2"]], R2_L_lmer = l[["R2"]],
         unique_M_lmer = a[["R2"]] - l[["R2"]], unique_L_lmer = a[["R2"]] - m[["R2"]], shared_lmer = m[["R2"]] + l[["R2"]] - a[["R2"]],
         any_singular = any(c(a[["singular"]], m[["singular"]], l[["singular"]]) == 1))
}

random_derangement <- function(k) { repeat { p <- sample.int(k); if (any(p != seq_len(k))) return(p) } }

analyse <- function(d, rows = seq_len(nrow(d$pairs)), lmer_fit = TRUE) {
  k <- length(d$cols); id <- seq_len(k)
  obs_long <- make_long(d, id, id, id, rows)
  obs <- stats_of(obs_long)
  perm_M_idx <- replicate(n_perm, random_derangement(k), simplify = FALSE)
  perm_L_idx <- replicate(n_perm, random_derangement(k), simplify = FALSE)
  boot_idx <- replicate(n_boot, sample.int(k, replace = TRUE), simplify = FALSE)
  run <- function(idx, f) do.call(rbind, parallel::mclapply(idx, f, mc.cores = n_cores))
  nullM <- run(perm_M_idx, function(p) stats_of(make_long(d, p, id, id, rows)))     # methylation from another colony
  nullL <- run(perm_L_idx, function(p) stats_of(make_long(d, p, id, p, rows)))      # lncRNA from another colony
  boot <- run(boot_idx, function(i) stats_of(make_long(d, i, i, i, rows)))
  q <- function(m, v, pr) unname(quantile(m[, v], pr, na.rm = TRUE))
  pval <- function(nl, v) (1 + sum(nl[, v] >= obs[[v]], na.rm = TRUE)) / (1 + sum(!is.na(nl[, v])))
  s <- tibble(species = d$species, pairs = length(rows), rows = obs[["rows"]], rows_path = obs[["rows_path"]],
              R2_combined = obs[["R2_combined"]], R2_M = obs[["R2_M"]], R2_L = obs[["R2_L"]],
              unique_M = obs[["unique_M"]], unique_M_lo = q(boot, "unique_M", 0.025), unique_M_hi = q(boot, "unique_M", 0.975),
              unique_M_null95 = q(nullM, "unique_M", 0.95), p_unique_M = pval(nullM, "unique_M"),
              unique_L = obs[["unique_L"]], unique_L_lo = q(boot, "unique_L", 0.025), unique_L_hi = q(boot, "unique_L", 0.975),
              unique_L_null95 = q(nullL, "unique_L", 0.95), p_unique_L = pval(nullL, "unique_L"),
              shared = obs[["shared"]], shared_lo = q(boot, "shared", 0.025), shared_hi = q(boot, "shared", 0.975),
              R2_combined_null95_M = q(nullM, "R2_combined", 0.95), R2_combined_null95_L = q(nullL, "R2_combined", 0.95),
              path_a = obs[["path_a"]], path_a_lo = q(boot, "path_a", 0.025), path_a_hi = q(boot, "path_a", 0.975),
              indirect = obs[["indirect"]], indirect_lo = q(boot, "indirect", 0.025), indirect_hi = q(boot, "indirect", 0.975),
              total_c = obs[["total_c"]], total_c_lo = q(boot, "total_c", 0.025), total_c_hi = q(boot, "total_c", 0.975),
              prop_mediated = obs[["prop_mediated"]])
  if (lmer_fit) s <- bind_cols(s, lmer_check(obs_long, d))
  list(summary = s, null = bind_rows(as_tibble(nullM) |> mutate(null = "methylation permuted"),
                                     as_tibble(nullL) |> mutate(null = "lncRNA permuted")) |> mutate(species = d$species))
}

classify <- function(s) {
  s |> mutate(
    M_pos = p_unique_M < 0.05, L_pos = p_unique_L < 0.05,
    no_signal = !M_pos & !L_pos,
    hierarchical = (path_a_lo > 0 | path_a_hi < 0) & (indirect_lo > 0 | indirect_hi < 0) &
                   sign(indirect) == sign(total_c) & prop_mediated >= 0.25,
    independent = M_pos & L_pos & shared < 0.25 * R2_combined,
    redundant = shared >= pmax(unique_M, unique_L) & R2_combined > R2_combined_null95_M & R2_combined > R2_combined_null95_L,
    pattern = case_when(no_signal ~ "no signal", hierarchical ~ "hierarchical", independent ~ "independent",
                        redundant ~ "redundant", TRUE ~ "unclassified"))
}

decide <- function(s) {
  s <- classify(s)
  n <- function(p) sum(s$pattern == p)
  verdict <- if (n("independent") >= 2) "supported"
    else if (n("redundant") >= 2 || n("hierarchical") >= 2) "not supported"
    else "inconclusive"
  list(per_species = s, verdict = verdict)
}
```

```{r fit}
res <- lapply(dat, analyse)
primary <- decide(bind_rows(lapply(res, `[[`, "summary")))
write_csv(primary$per_species, file.path(out_dir, "commonality_summary.csv"))
write_csv(bind_rows(lapply(res, `[[`, "null")), file.path(out_dir, "null_draws.csv"))
knitr::kable(primary$per_species |> select(species, pairs, rows, R2_combined, unique_M, p_unique_M, unique_L, p_unique_L,
                                           shared, pattern), digits = 5)
```

**Primary verdict: `r primary$verdict`.**

Hierarchical path (lncRNA-locus methylation → lncRNA → gene):

```{r path}
knitr::kable(primary$per_species |> select(species, rows_path, path_a, path_a_lo, path_a_hi, indirect, indirect_lo, indirect_hi,
                                           total_c, total_c_lo, total_c_hi, prop_mediated, hierarchical), digits = 4)
```

OLS vs the pre-registered `lmer` fits on the observed data:

```{r lmer-check}
chk <- primary$per_species |> transmute(species,
  d_R2_combined = R2_combined - R2_combined_lmer, d_unique_M = unique_M - unique_M_lmer,
  d_unique_L = unique_L - unique_L_lmer, d_shared = shared - shared_lmer, any_singular)
chk <- chk |> mutate(max_abs = pmax(abs(d_R2_combined), abs(d_unique_M), abs(d_unique_L), abs(d_shared)),
                     within_tolerance = max_abs < 1e-4)
write_csv(chk, file.path(out_dir, "ols_vs_lmer.csv"))
knitr::kable(chk, digits = 7)
```

```{r fig, fig.width = 9, fig.height = 3.8}
fig_dat <- primary$per_species |>
  transmute(species, `unique methylation` = unique_M, `unique lncRNA` = unique_L, shared,
            M_lo = unique_M_lo, M_hi = unique_M_hi, L_lo = unique_L_lo, L_hi = unique_L_hi, S_lo = shared_lo, S_hi = shared_hi) |>
  pivot_longer(c(`unique methylation`, `unique lncRNA`, shared), names_to = "component", values_to = "R2") |>
  mutate(lo = case_when(component == "unique methylation" ~ M_lo, component == "unique lncRNA" ~ L_lo, TRUE ~ S_lo),
         hi = case_when(component == "unique methylation" ~ M_hi, component == "unique lncRNA" ~ L_hi, TRUE ~ S_hi)) |>
  select(species, component, R2, lo, hi)
write_csv(fig_dat, file.path(out_dir, "commonality_figure_data.csv"))
g <- ggplot(fig_dat, aes(component, R2)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_pointrange(aes(ymin = lo, ymax = hi), colour = "#1b6ca8") +
  facet_wrap(~ species) +
  labs(x = NULL, y = "Within-colony expression variance (R2)") +
  theme_bw() + theme(axis.text.x = element_text(angle = 20, hjust = 1))
ggsave(file.path(out_dir, "commonality.png"), g, width = 9, height = 3.8, dpi = 150)
g
```

Points are commonality fractions; bars are 95% colony-bootstrap intervals.

## Sensitivity check: genes with >= 20 CpGs

```{r sens}
res20 <- lapply(dat, function(d) analyse(d, which(d$pairs$n_cpg >= 20), lmer_fit = FALSE))
sens20 <- decide(bind_rows(lapply(res20, `[[`, "summary")))
write_csv(sens20$per_species, file.path(out_dir, "sens_mincpg20_summary.csv"))
knitr::kable(sens20$per_species |> select(species, pairs, R2_combined, unique_M, p_unique_M, unique_L, p_unique_L, shared, pattern), digits = 5)
```

```{r summary}
verdicts <- tibble(analysis = c("primary (>= 5 CpGs)", ">= 20 CpGs"), verdict = c(primary$verdict, sens20$verdict))
write_csv(verdicts, file.path(out_dir, "verdicts.csv"))
knitr::kable(verdicts)
```

## Caveats

- Samples with all three layers: `r paste(n_tab$species, n_tab$samples, sep = " ", collapse = ", ")`;
  about 10 colonies per species.
- Within-colony variation over at most four timepoints per colony.
- lncRNA strand is unknown (D-014), and H17 found most cis coupling to be
  generic local co-expression, so unique_L is "nearby lncRNA", not
  "regulatory lncRNA".
- miRNA, a third channel, is out of scope (no target predictions in the
  manifest).
- CpG density differs about 20-fold across species (D-008).
