---
title: "H19: methylome turnover and energy reserves"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
root <- tryCatch(rprojroot::find_root(rprojroot::has_file("_targets.R")), error = function(e) "../..")
knitr::opts_knit$set(root.dir = root)
setwd(root)
source(file.path(root, "R", "load.R"))
source(file.path(root, "R", "expression.R"))
suppressPackageStartupMessages({ library(tidyr); library(ggplot2) })
design <- load_design()
out_dir <- file.path(root, "hypotheses", "H19", "output")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species_all <- c("Apul", "Peve", "Ptuh")
n_boot <- 2000
cov_floor <- 10
set.seed(20261008)
```

## Pre-registered decision rule

> - `supported` if the pooled slope is negative with 95% CI excluding 0 and
>   the per-species estimate is negative in at least two species.
> - `not supported` if the pooled slope is >= 0, or the CI includes 0 and
>   |slope| < 0.2 SD.
> - `inconclusive` otherwise.

Amendment 2 (after the first run showed the median-over-all-CpGs score is
degenerate; written before any association was seen) restricts turnover to
CpGs with species-mean methylation 10–90% and uses the mean over them.

Pooled slope: `lm(z(AFDW_mean) ~ z(turnover_meth) + species)`, z within
species, one row per colony with CpG data at >= 3 timepoints. Per-species
estimate: Spearman rho (Amendment 1). Amendment 1 (`hypothesis.md`, before
any analysis) also fixes the noise correction at the 10x coverage floor
because the CpG layer has no coverage (D-015), and defines the lipid ratio
as log((TG + DG + CE) / (LPC + Cer)).

Implementation notes (not in `hypothesis.md`; none changes the rule):

- Methylation turnover uses all of a colony's CpG timepoints (3 or 4);
  expression turnover uses the same timepoints where expression exists.
- Colony means of physiology, lipid ratio and log(Methionine/SAM) are taken
  over all of that colony's samples in the layer, at any timepoint.
- The rule is applied as written to the primary model. Secondary and
  sensitivity models are reported with the same slope, CI and rho columns;
  verdicts for sensitivity models use the same rule.

## Data

```{r turnover}
turnover_species <- function(sp) {
  cpg <- as_count_matrix(load_cpg(sp)) / 100
  mu <- rowMeans(cpg)
  cpg <- cpg[mu >= 0.1 & mu <= 0.9, , drop = FALSE]   # Amendment 2: CpGs that can vary
  smp <- design |> filter(species == sp, has_cpg %in% TRUE, sample_id %in% colnames(cpg))
  keep_cols <- smp |> count(colony) |> filter(n >= 3) |> pull(colony)
  ex <- prep_vst(load_genes(sp), design)
  one <- function(col, tps = NULL) {
    s <- smp |> filter(colony == col, is.null(tps) | timepoint %in% tps)
    if (nrow(s) < 3) return(NULL)
    p <- cpg[, s$sample_id, drop = FALSE]
    raw <- matrixStats::rowVars(p)
    noise <- rowMeans(p * (1 - p)) / cov_floor
    e_ids <- intersect(s$sample_id, colnames(ex$vst))
    tibble(species = sp, colony = col, n_tp = nrow(s),
           turnover_meth = mean(raw - noise), turnover_meth_median = median(raw - noise),
           turnover_meth_raw = mean(raw),
           turnover_expr = if (length(e_ids) >= 3) median(matrixStats::rowVars(ex$vst[, e_ids, drop = FALSE])) else NA_real_,
           n_tp_expr = length(e_ids))
  }
  list(all = bind_rows(lapply(keep_cols, one)),
       tp134 = bind_rows(lapply(keep_cols, one, tps = c("TP1", "TP3", "TP4"))),
       n_cpg = nrow(cpg), n_cpg_all = length(mu))
}
tv <- lapply(setNames(species_all, species_all), turnover_species)
turn <- bind_rows(lapply(tv, `[[`, "all"))
turn134 <- bind_rows(lapply(tv, `[[`, "tp134"))
```

```{r phenotypes}
colony_info <- design |> distinct(colony, species, site, nutrient)
phys <- load_physiology() |> select(sample_id, Host_AFDW.mg.cm2, prot_mg.cm2, cells.cm2) |>
  left_join(design |> select(sample_id, colony), by = "sample_id") |>
  group_by(colony) |>
  summarise(AFDW = mean(Host_AFDW.mg.cm2, na.rm = TRUE),
            AFDW_per_prot = mean(Host_AFDW.mg.cm2 / prot_mg.cm2, na.rm = TRUE),
            cells = mean(cells.cm2, na.rm = TRUE), n_phys = n(), .groups = "drop")

lip <- load_lipidomics()
lclass <- sub("[ (].*", "", lip$feature_id)
lipid_map <- tibble(feature_id = lip$feature_id, class = lclass,
                    group = case_when(lclass %in% c("TG", "DG", "CE") ~ "storage",
                                      lclass %in% c("LPC", "Cer") ~ "membrane", TRUE ~ "neither"))
write_csv(lipid_map, file.path(out_dir, "lipid_class_map.csv"))
lm_ <- as_count_matrix(lip)
ratio <- tibble(sample_id = colnames(lm_),
                storage = colSums(lm_[lipid_map$group == "storage", , drop = FALSE], na.rm = TRUE),
                membrane = colSums(lm_[lipid_map$group == "membrane", , drop = FALSE], na.rm = TRUE)) |>
  mutate(log_ratio = log(storage / membrane)) |>
  left_join(design |> select(sample_id, colony), by = "sample_id") |>
  group_by(colony) |> summarise(lipid_ratio = mean(log_ratio[is.finite(log_ratio)]), n_lipid = n(), .groups = "drop")

met <- load_metabolomics()
mm <- as_count_matrix(met)
metab <- tibble(sample_id = colnames(mm), met = mm["Methionine", ], sam = mm["S-Adenosylmethionine (SAM)", ]) |>
  mutate(log_met_sam = log(met / sam)) |>
  left_join(design |> select(sample_id, colony), by = "sample_id") |>
  group_by(colony) |> summarise(log_met_sam = mean(log_met_sam[is.finite(log_met_sam)]), n_metab = n(), .groups = "drop")

dat <- turn |> left_join(colony_info, by = c("colony", "species")) |>
  left_join(phys, by = "colony") |> left_join(ratio, by = "colony") |> left_join(metab, by = "colony")
write_csv(dat, file.path(out_dir, "colony_table.csv"))

n_tab <- dat |> group_by(species) |>
  summarise(colonies = n(), with_4_tp = sum(n_tp == 4), with_afdw = sum(is.finite(AFDW)),
            with_lipid = sum(is.finite(lipid_ratio)), with_met_sam = sum(is.finite(log_met_sam)), .groups = "drop") |>
  left_join(tibble(species = species_all, cpgs_all = sapply(tv, `[[`, "n_cpg_all"), cpgs_10_90 = sapply(tv, `[[`, "n_cpg")), by = "species")
write_csv(n_tab, file.path(out_dir, "n_per_species.csv"))
knitr::kable(n_tab, format.args = list(big.mark = ","))
```

Lipid classes (`lipid_class_map.csv`):

```{r lipid-map}
knitr::kable(lipid_map |> count(class, group))
```

## Model

```{r functions}
zs <- function(x) if (sum(is.finite(x)) > 1 && sd(x, na.rm = TRUE) > 0) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE) else x * NA

fit_one <- function(d, response, predictor = "turnover_meth", covars = character()) {
  d <- d |> filter(is.finite(.data[[response]]), is.finite(.data[[predictor]]),
                   if (length(covars)) if_all(all_of(covars), ~ !is.na(.x)) else TRUE) |>
    group_by(species) |> filter(n() >= 3) |>
    mutate(y = zs(.data[[response]]), x = zs(.data[[predictor]]),
           across(all_of(covars), ~ if (is.numeric(.x)) zs(.x) else .x, .names = "c_{.col}")) |> ungroup()
  rhs <- paste(c("x", "species", if (length(covars)) paste0("c_", covars)), collapse = " + ")
  f <- lm(as.formula(paste("y ~", rhs)), data = d)
  ci <- confint(f)["x", ]
  rho <- d |> group_by(species) |> group_modify(function(g, k) {
    b <- replicate(n_boot, { i <- sample.int(nrow(g), replace = TRUE); suppressWarnings(cor(g$x[i], g$y[i], method = "spearman")) })
    tibble(n = nrow(g), rho = cor(g$x, g$y, method = "spearman"),
           rho_lo = quantile(b, 0.025, na.rm = TRUE), rho_hi = quantile(b, 0.975, na.rm = TRUE))
  }) |> ungroup()
  list(fit = tibble(response = response, predictor = predictor, covariates = paste(covars, collapse = "+"),
                    n = nrow(d), slope = unname(coef(f)["x"]), lo = ci[1], hi = ci[2],
                    p = summary(f)$coefficients["x", 4],
                    n_rho_negative = sum(rho$rho < 0, na.rm = TRUE)),
       rho = rho |> mutate(response = response, covariates = paste(covars, collapse = "+"), .before = 1),
       data = d)
}

decide <- function(r) {
  with(r, if (slope < 0 && hi < 0 && n_rho_negative >= 2) "supported"
          else if (slope >= 0 || (lo <= 0 && hi >= 0 && abs(slope) < 0.2)) "not supported"
          else "inconclusive")
}
```

```{r primary}
prim <- fit_one(dat, "AFDW")
primary_verdict <- decide(prim$fit)
write_csv(prim$fit |> mutate(verdict = primary_verdict), file.path(out_dir, "primary.csv"))
write_csv(prim$rho, file.path(out_dir, "primary_per_species_rho.csv"))
knitr::kable(prim$fit, digits = 3)
knitr::kable(prim$rho, digits = 3)
```

**Primary verdict: `r primary_verdict`.**

```{r fig, fig.width = 9, fig.height = 3.6}
fd <- prim$data |> select(species, colony, x, y)
write_csv(fd, file.path(out_dir, "figure_data.csv"))
g <- ggplot(fd, aes(x, y)) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey60", linewidth = 0.6) +
  geom_point(colour = "#1b6ca8", size = 2) + facet_wrap(~ species) +
  labs(x = "Excess methylation turnover (z within species)", y = "Host AFDW, colony mean (z within species)") +
  theme_bw()
ggsave(file.path(out_dir, "afdw_vs_turnover.png"), g, width = 9, height = 3.6, dpi = 150)
g
```

## Secondary (BH across the three)

```{r secondary}
sec <- list(fit_one(dat, "lipid_ratio"), fit_one(dat, "log_met_sam"),
            fit_one(dat, "AFDW", covars = "turnover_expr"))
sec_tab <- bind_rows(lapply(sec, `[[`, "fit")) |> mutate(p_bh = p.adjust(p, "BH"))
write_csv(sec_tab, file.path(out_dir, "secondary.csv"))
write_csv(bind_rows(lapply(sec, `[[`, "rho")), file.path(out_dir, "secondary_per_species_rho.csv"))
knitr::kable(sec_tab, digits = 3)
```

Predicted signs: lipid ratio negative, log(Methionine/SAM) positive, AFDW
negative after adjusting for expression turnover.

## Sensitivity checks

```{r sensitivity}
dat134 <- turn134 |> left_join(colony_info, by = c("colony", "species")) |> left_join(phys, by = "colony")
sens <- list(
  `TP1, TP3, TP4 only` = fit_one(dat134, "AFDW"),
  `site covariate` = fit_one(dat, "AFDW", covars = "site"),
  `AFDW per protein` = fit_one(dat, "AFDW_per_prot"),
  `uncorrected variance` = fit_one(dat, "AFDW", predictor = "turnover_meth_raw"),
  `median over CpGs` = fit_one(dat, "AFDW", predictor = "turnover_meth_median"),
  `cells.cm2 covariate` = fit_one(dat, "AFDW", covars = "cells"))
sens_tab <- bind_rows(lapply(names(sens), function(k) sens[[k]]$fit |> mutate(check = k, .before = 1))) |>
  rowwise() |> mutate(verdict = decide(pick(everything()))) |> ungroup()
write_csv(sens_tab, file.path(out_dir, "sensitivity.csv"))
knitr::kable(sens_tab, digits = 3)
```

Rank agreement of corrected and uncorrected turnover within species:

```{r rank-agree}
ra <- dat |> group_by(species) |> summarise(spearman_corrected_vs_raw = cor(turnover_meth, turnover_meth_raw, method = "spearman"), .groups = "drop")
knitr::kable(ra, digits = 3)
```

## Descriptive: species means (no ranking claim)

```{r species-means}
sm <- dat |> group_by(species) |> summarise(across(c(turnover_meth, turnover_expr, AFDW, lipid_ratio, log_met_sam), ~ mean(.x, na.rm = TRUE)), .groups = "drop")
write_csv(sm, file.path(out_dir, "species_means.csv"))
knitr::kable(sm, digits = 4)
```

```{r summary}
verdicts <- tibble(analysis = c("primary", sens_tab$check), verdict = c(primary_verdict, sens_tab$verdict))
write_csv(verdicts, file.path(out_dir, "verdicts.csv"))
knitr::kable(verdicts)
```

## Caveats

- About 10 colonies per species, pooled n in the table above; the 95% CI
  half-width on a standardized slope at n about 27 is about 0.4 SD.
- No per-sample coverage (D-015): the noise correction is a fixed 10x bound
  and colony differences in sequencing depth are not removed.
- CpG sets differ about 20-fold across species (D-008); turnover is
  standardized within species before pooling.
